---
title: "How Do Human-Modified Landscapes Shape Bat Observations in California?"
subtitle: "Environmental Data Science Capstone Project"
author: "Michele Perry"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: cosmo
  github_document:
    toc: true
    html_preview: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center"
)

options(scipen = 999, dplyr.summarise.inform = FALSE)
set.seed(123)
```

## 1. Introduction

Urbanization reshapes landscapes at every scale — replacing habitat with
impervious surface, flooding the night sky with artificial light, and
concentrating human activity into dense corridors. For wildlife, these
changes create winners and losers. Some species thrive at the urban
edge; others retreat to darker, less disturbed landscapes. Understanding
which species fall into which category is central to managing
biodiversity in an increasingly human-dominated world.

Bats are a particularly informative group for studying these dynamics.
As nocturnal insectivores, they are directly affected by artificial
light (which can disrupt foraging behavior and prey availability), land
development (which removes roosting habitat), and the broader footprint
of urbanization. California, with its steep urban-to-wildland gradients
and rich bat fauna, provides an ideal setting to ask how these pressures
structure species distributions.

This project examines three focal bat species with contrasting
ecological profiles:

-   *Aeorestes cinereus* (hoary bat) — a long-distance migratory,
    tree-roosting species
-   *Myotis yumanensis* (Yuma myotis) — a water-associated species often
    found near bridges and buildings
-   *Myotis californicus* (California myotis) — a small,
    crevice-roosting species widespread across the state

Using publicly available occurrence records from the [Global
Biodiversity Information Facility (GBIF)](https://www.gbif.org/), this
study asks:

> **How do components of human-modified landscapes structure bat
> observation patterns at fine spatial scales, and do species respond
> differently to increasing urban intensity?**

### A critical framing note: observer bias in opportunistic data

GBIF records are opportunistic — they reflect where people *looked*, not
just where bats *are*. Observations cluster near roads, cities, and
research stations, meaning that any apparent association between bats
and urbanization could partly reflect reporting effort rather than true
ecological preference. This project addresses this bias explicitly by
including population density as an accessibility proxy and by
interpreting results as patterns in *observed* presence rather than
confirmed ecological effects.

## 2. Data acquisition and processing

The data processing pipeline lives in the `R/` directory as a series of
modular scripts (`R/00_setup.R` through
`R/06_build_grid_model_dataset.R`) that can be run in sequence to
reproduce the full workflow from raw downloads to the final
analysis-ready dataset. This section walks through each stage in the
order the data were actually acquired and processed, highlighting key
spatial operations and design decisions along the way.

```{r load-libraries}
library(tidyverse)
library(sf)
library(scales)
library(viridis)
library(here)
library(broom)
library(pROC)
library(rsample)
library(purrr)
```

### 2.1 GBIF bat occurrence data

The starting point for this project is publicly available species
occurrence data from the [Global Biodiversity Information Facility
(GBIF)](https://www.gbif.org/). GBIF aggregates biodiversity records
from museum collections, citizen science platforms, and research surveys
into a single downloadable format.

The `rgbif` package provides an R interface to the GBIF API. The
workflow in `R/01_get_clean_bat_points.R` begins by resolving each focal
species name against the GBIF backbone taxonomy using `name_backbone()`,
which returns a unique `species_key` — a stable numeric identifier that
avoids issues with taxonomic synonyms.

The download request is submitted via `occ_download()` with filters for
the three focal species keys, valid coordinates, and years 2012–2024:

```{r gbif-download-example, eval=FALSE}
# Submitted via rgbif (not re-run here; see R/01_get_clean_bat_points.R)
dl <- occ_download(
  pred_in("taxonKey", taxa_keys),
  pred("hasCoordinate", TRUE),
  pred_gte("year", 2012),
  pred_lte("year", 2024),
  format = "SIMPLE_CSV"
)
```

GBIF processes this request server-side and returns a zipped CSV with
dozens of columns. The cleaning step selects only the fields needed for
this analysis and applies coordinate hygiene filters:

```{r cleaning-example, eval=FALSE}
# Simplified from R/01_get_clean_bat_points.R
dat <- raw %>%
  transmute(
    gbif_id, species, scientific_name,
    year, month, day,
    lon = decimal_longitude,
    lat = decimal_latitude,
    coordinate_uncertainty_m = coordinate_uncertainty_in_meters
  ) %>%
  filter(
    !is.na(lon), !is.na(lat),        # must have coordinates
    lon != 0, lat != 0,               # exclude zero-island artifacts
    lon >= -125, lon <= -113,          # rough CA bounding box
    lat >= 32, lat <= 42
  )
```

After filtering, the cleaned records are converted to an `sf` spatial
points object in WGS 84 (EPSG:4326) using `st_as_sf()` — this is the
starting point for all downstream spatial operations.

### 2.2 California state boundary

A California state boundary polygon is needed both for spatially
filtering bat points and for constructing the analysis grid.
`R/02_get_ca_boundary.R` retrieves this from the US Census TIGER/Line
database using the `tigris` package:

```{r boundary-example, eval=FALSE}
# From R/02_get_ca_boundary.R
ca <- tigris::states(cb = TRUE, year = 2022) %>%
  filter(STUSPS == "CA") %>%
  st_make_valid()
```

Let's load the boundary and the cleaned bat points to see what we're
working with:

```{r load-boundary-and-points}
ca <- st_read(
  here("data", "processed", "boundaries", "ca_boundary.gpkg"),
  layer = "ca_boundary", quiet = TRUE
)

pts <- st_read(
  here("data", "processed", "gbif", "gbif_bats_points_clean_CA_2012_2024.gpkg"),
  quiet = TRUE
)
```

```{r points-summary}
cat("Total bat occurrence records in California:", nrow(pts), "\n")
pts %>% st_drop_geometry() %>% count(species, sort = TRUE)
```

These 455 records are the result of filtering the raw GBIF download to
valid California coordinates and then performing a spatial intersection
with the state boundary (`st_intersects()` in
`R/03_build_grid_and_domain.R`).

### 2.3 Spatial framework: 10 km equal-area grid

With the bat points and boundary in hand, the next decision is how to
structure the analysis spatially. Rather than using administrative units
like counties (which vary enormously in size and span multiple habitat
types), this project uses a regular 10 km equal-area grid. A uniform
grid ensures that every analysis unit covers the same area, making
landscape summaries directly comparable across cells.

All spatial operations use the California Albers projection (EPSG:3310),
an equal-area coordinate system measured in meters — essential for any
analysis where area calculations matter.

The grid is constructed in `R/03_build_grid_and_domain.R` using
`sf::st_make_grid()`:

```{r grid-construction-example, eval=FALSE}
# From R/03_build_grid_and_domain.R
grid_full <- st_make_grid(
  ca_ae,                     # CA boundary in Albers projection
  cellsize = 10000,          # 10 km cells
  square = TRUE
) %>%
  st_as_sf() %>%
  mutate(cell_id = row_number())
```

This produces a grid of regular squares covering the full bounding box
of California. The analysis grid is then trimmed using two criteria: (1)
cells whose centroids fall inside California, and (2) a "presence
rescue" step that retains any cell containing at least one bat
observation, even if its centroid falls just outside the boundary. This
prevents edge-case data loss along the coastline and state borders.

**Defining the accessible domain.** Not all of California has been
surveyed for bats. To avoid treating genuinely unsampled areas as
ecological absences, the analysis is restricted to an "accessible"
subset of the grid. This domain is defined by buffering every bat
observation point by 50 km using `st_buffer()`, taking the union of all
buffers, and intersecting the result with a buffered California
boundary. Grid cells with centroids inside this domain (plus
presence-rescued cells) form the final accessible grid:

```{r load-grid}
grid <- st_read(
  here("data", "processed", "grid", "ca_grid10km_accessible.gpkg"),
  quiet = TRUE
)

cat("Total accessible grid cells:", nrow(grid), "\n")
```

```{r map-grid-and-points, fig.width=8, fig.height=9}
pts_map <- st_transform(pts, st_crs(grid))
ca_map  <- st_transform(ca, st_crs(grid))

ggplot() +
  geom_sf(data = grid, fill = "gray90", color = "white", linewidth = 0.2) +
  geom_sf(data = ca_map, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts_map, color = "#2CBEC9", size = 1.5, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "10 km accessible grid with bat observation locations",
    subtitle = "2,811 grid cells covering the accessible modeling domain"
  ) +
  theme_minimal(base_size = 13)
```

The grid covers the portions of California where bat observations have
been recorded (plus a 50 km buffer), excluding remote areas like the far
eastern deserts and high Sierra peaks where no records exist. This
represents roughly 75% of the total California grid.

### 2.4 Landscape covariates

Three landscape variables were computed for each grid cell, each drawn
from a different remote sensing or geospatial dataset. The processing
pipeline (`R/04_build_grid_covariates.R`) involves reprojecting rasters
and vectors to match the grid CRS, cropping and masking to the study
extent, and extracting summary statistics per cell.

#### VIIRS nighttime light radiance

The Visible Infrared Imaging Radiometer Suite (VIIRS) produces annual
composite images of nighttime light intensity. These rasters measure
radiance in nW/cm²/sr — brighter values indicate more artificial light.
Annual composites for 2012–2024 were downloaded from the Earth
Observation Group (VNL v2.1 for 2012–2021 and v2.2 for 2022–2024).

Each annual GeoTIFF was cropped to the grid extent, reprojected from its
native WGS 84 to California Albers, and then summarized per grid cell
using `exactextractr::exact_extract()` with an area-weighted mean. The
per-year means were then averaged across all available years to produce
a single long-term mean radiance value per cell:

```{r viirs-example, eval=FALSE}
# Simplified from R/04_build_grid_covariates.R
r <- terra::rast(tif_path)
r_crop <- terra::crop(r, grid_extent) |> terra::mask(grid_extent)
r_ae <- terra::project(r_crop, "EPSG:3310", method = "bilinear")

cell_means <- exactextractr::exact_extract(r_ae, grid, "mean")
```

#### NLCD percent developed land

The National Land Cover Database (NLCD) 2019 is a 30-meter categorical
raster classifying every pixel in the contiguous US into land cover
types. The relevant categories for this analysis are the four
"Developed" classes: Open Space, Low Intensity, Medium Intensity, and
High Intensity.

To compute percent developed land per grid cell, the NLCD raster was
cropped to the grid extent and extracted using `terra::extract()` with
`fun = table`, which counts the number of pixels in each land cover
class within each cell. The proportion of developed pixels was then
calculated:

```{r nlcd-example, eval=FALSE}
# Simplified from R/04_build_grid_covariates.R
nlcd_crop <- terra::crop(nlcd, grid_extent) |> terra::mask(grid_extent)
ex_tab <- terra::extract(nlcd_crop, grid_polygons, fun = table, na.rm = TRUE)

# Sum all "Developed*" class columns, divide by total pixels
pct_developed <- dev_cells / total_cells
```

#### GPWv4 population density

The Gridded Population of the World, Version 4 (GPWv4, Revision 11) from
NASA SEDAC provides population density estimates at \~1 km resolution.
Like the VIIRS rasters, this was cropped, reprojected to California
Albers, and summarized per grid cell using an area-weighted mean via
`exact_extract()`.

#### PAD-US percent protected land

A fourth covariate — percent protected land — was also computed from
the Protected Areas Database of the United States (PAD-US 4.1), using a
spatial intersection (`st_intersection()`) between grid cells and
protected area polygons with GAP Status 1–3. However, exploratory
analysis showed that protected land coverage added minimal predictive
value for bat presence once the urban variables were included, so it is
not featured in the final modeling. The processing code is preserved in
`R/04_build_grid_covariates.R` for reference.

#### Combined covariate dataset

All three urban covariates were joined by `cell_id` into a single
table. Log-transformed versions of radiance and population density
(`log1p_radiance`, `log1p_pop_density`) were computed to reduce right
skew — both variables span several orders of magnitude, with most cells
near zero and a long tail of urban cells with very high values.

```{r load-covariates}
covariates <- read_csv(
  here("data", "processed", "covariates_grid", "grid_covariates_10km.csv"),
  show_col_types = FALSE
)

glimpse(covariates)
```

```{r covariate-summary}
covariates %>%
  summarise(
    n_cells = n(),
    across(c(mean_radiance, pct_developed, pop_density),
           list(min = ~min(., na.rm = TRUE),
                median = ~median(., na.rm = TRUE),
                max = ~max(., na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value")
```

The ranges confirm what we'd expect: most cells have very low radiance,
little developed land, and sparse population, but a small number of
urban cells have extreme values across all three variables.

### 2.5 Building the final analysis dataset

The last step brings together the species presence data and the
landscape covariates into a single analysis-ready dataset.

**Why aggregate across years?** The landscape covariates in this study
— nighttime light, land development, population density — vary far more
across space than over time. A rural cell in the Sierra foothills and an
urban cell in Los Angeles differ by orders of magnitude in radiance and
population density, but neither changes much from 2012 to 2024.
Variation in these predictors is overwhelmingly spatial, not temporal.
On the response side, 455 total observations spread across 13 years and
three species leaves far too few records per year to support meaningful
year-level modeling. Aggregating across the full study period pools all
available data into a single spatial snapshot, maximizing statistical
power at the cost of collapsing any temporal dynamics — a tradeoff
revisited in the limitations section.

**Presence aggregation.** Each bat point was spatially joined to the
accessible grid using `st_join()` with `st_within()` — assigning every
observation to the grid cell it falls inside. Observations were then
aggregated by cell and species across all years (2012–2024).

To create a complete panel — including cells where a species was *not*
observed — `tidyr::complete()` was used to fill in every cell × species
combination with explicit zeros. A binary `is_present` column was
derived as `as.integer(n_obs > 0)`:

```{r presence-example, eval=FALSE}
# Simplified from R/05_build_grid_presence.R
agg <- pts_joined %>%
  group_by(cell_id, species) %>%
  summarise(n_obs = n_distinct(gbif_id))

presence <- agg %>%
  complete(
    cell_id = grid$cell_id,
    species = species_levels,
    fill = list(n_obs = 0L)
  ) %>%
  mutate(is_present = as.integer(n_obs > 0))
```

**Final merge.** The presence panel was joined to the covariate table by
`cell_id` using a left join, producing one row per cell × species with
both response and predictor variables:

```{r load-final-dataset}
df <- read_csv(
  here("data", "processed", "analysis_grid", "grid_model_dataset_10km.csv"),
  show_col_types = FALSE
)
```

```{r final-dataset-glimpse}
glimpse(df)
```

```{r final-dataset-summary}
df %>%
  summarise(
    n_rows = n(),
    n_cells = n_distinct(cell_id),
    n_species = n_distinct(species),
    total_observations = sum(n_obs),
    overall_presence_rate = round(mean(is_present), 4)
  )
```

The final dataset contains 8,433 rows (2,811 cells × 3 species), each
with three landscape covariates and their log-transformed versions. This
is the dataset used for all exploration and modeling that follows.

## 3. Exploratory data analysis

### 3.1 Response structure: extreme sparsity

```{r presence-absence-bar}
pres_summary <- df %>%
  group_by(species) %>%
  summarise(
    n_cells = n(),
    n_present = sum(is_present),
    prop_present = round(mean(is_present), 3),
    .groups = "drop"
  )

df %>%
  mutate(
    status = if_else(is_present == 1, "Present (\u22651 record)", "Absent (0 records)")
  ) %>%
  count(species, status) %>%
  group_by(species) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = species, y = prop, fill = status)) +
  geom_col() +
  geom_text(
    data = pres_summary,
    aes(x = species, y = prop_present, label = prop_present),
    inherit.aes = FALSE,
    hjust = -0.3, size = 5, fontface = "bold", color = "#1D9E75"
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Present (\u22651 record)" = "#1D9E75",
               "Absent (0 records)" = "gray75")
  ) +
  labs(
    title = "Most grid cells have no bat observations",
    subtitle = "Proportion of 2,811 accessible cells with at least one record (2012\u20132024)",
    x = NULL,
    y = "Proportion of cells",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

All three species are observed in fewer than 5% of grid cells.
*Aeorestes cinereus* is the most frequently observed, while
*Myotis californicus* is the rarest. This extreme sparsity
motivates modeling binary presence (observed or not) rather than
observation counts — any count-based model would be overwhelmed by
structural zeros.

### 3.2 Mapping observations over the human-modified landscape

```{r prepare-map-data}
# Join covariates to the spatial grid for mapping
grid_map <- grid %>%
  mutate(cell_id = as.character(cell_id)) %>%
  left_join(
    df %>%
      mutate(cell_id = as.character(cell_id)) %>%
      distinct(cell_id, log1p_pop_density, log1p_radiance, pct_developed),
    by = "cell_id"
  )

# Ensure CRS alignment for all layers
pts_map <- st_transform(pts, st_crs(grid_map))
ca_map  <- st_transform(ca, st_crs(grid_map))
```

```{r map-pop-density, fig.width=8, fig.height=9}
# Define breaks in original units (roughly log-spaced)
pop_breaks <- c(0, 1, 10, 100, 1000, 5000, Inf)
pop_labels <- c("0 – 1", "1 – 10", "10 – 100", "100 – 1,000",
                "1,000 – 5,000", "> 5,000")

# Bin the log-transformed values using the transformed breaks
grid_map <- grid_map %>%
  mutate(
    pop_bin = cut(
      log1p_pop_density,
      breaks = log1p(pop_breaks),
      labels = pop_labels,
      include.lowest = TRUE,
      right = FALSE
    )
  )

# Viridis colors sampled at 6 evenly spaced points
bin_colors <- viridis::viridis(6, option = "cividis")

ggplot() +
  geom_sf(data = grid_map, aes(fill = pop_bin), color = NA) +
  scale_fill_manual(
    values = setNames(bin_colors, pop_labels),
    name = "Population density\n(people/km\u00B2)",
    na.value = "gray90",
    drop = FALSE
  ) +
  geom_sf(data = ca_map, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts_map, color = "black", fill = "#2CBEC9",
          shape = 21, size = 2.5, stroke = 0.6, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "GBIF bat observations cluster near urban areas",
    subtitle = "Cyan points = bat observations; background = population density on accessible 10 km grid"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.5, "cm")
  )
```

The map reveals a clear pattern: bat observations are concentrated in
and around California's major population centers — the San Francisco Bay
Area, Los Angeles basin, San Diego, and Sacramento. Large swaths of the
state's interior, mountains, and deserts have few or no records. This is
the observer bias problem in action: people (and therefore observation
effort) concentrate in cities.

Maps of nighttime light radiance and percent developed land (not shown)
display nearly identical spatial patterns — the same urban cores light
up, the same rural areas stay dark. This visual redundancy is confirmed
by the correlation analysis below.

### 3.3 Predictor correlations

```{r correlation-matrix}
cov_df <- df %>%
  select(log1p_radiance, log1p_pop_density, pct_developed) %>%
  drop_na()

cor_mat <- round(cor(cov_df), 3)
cor_mat
```

Nighttime light, population density, and percent developed land are all
strongly correlated (r = 0.76–0.87). At 10 km resolution, these
variables essentially capture the same underlying gradient of human
modification. Including all three in a single model would introduce
redundancy without adding predictive value — a finding that the model
comparison in the next section confirms formally.

### 3.4 Presence rates along the population density gradient

To explore how bat presence varies with urban intensity, we divide
population density into decile bins and compute the share of cells with
at least one observation in each bin, separately by species.

```{r species-colors}
# Consistent species color palette used throughout
species_colors <- c(
  "Aeorestes cinereus"  = "#1B9E77",
  "Myotis yumanensis"   = "#D95F02",
  "Myotis californicus" = "#7570B3"
)
```

```{r presence-vs-pop, fig.height=5}
df %>%
  mutate(bin = ntile(log1p_pop_density, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_pop_density, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = species_colors) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Presence rate increases with population density",
    subtitle = "Binned deciles of log(1 + population density)",
    x = "Mean log(1 + population density) within bin",
    y = "Presence rate",
    color = "Species"
  )
```

Presence rates rise sharply in the upper deciles of population density,
and the three species diverge: *Aeorestes cinereus* shows the steepest
increase, while *Myotis californicus* responds more weakly. Equivalent
plots for nighttime radiance and percent developed land (not shown)
display the same qualitative pattern — unsurprising given the high
correlations among these predictors. These exploratory patterns motivate
the formal modeling that follows.

## 4. Model comparison: what predicts bat presence?

Given the strong collinearity among urban predictors, a natural question
is: **does adding nighttime light or land development improve
predictions beyond population density alone?** This matters both
practically (simpler models are easier to interpret) and conceptually
(if the variables are interchangeable, they're measuring the same
thing).

### 4.1 Cross-validated model comparison

We fit a series of logistic regression models predicting binary presence
(`is_present`) and compare their performance using 5-fold
cross-validated AUC (area under the ROC curve). Models are fit
separately for each species to allow species-specific effects.

```{r standardize-predictors}
# Standardize predictors for comparable coefficients
df_model <- df %>%
  mutate(
    z_radiance  = scale(log1p_radiance)[, 1],
    z_pop       = scale(log1p_pop_density)[, 1],
    z_dev       = scale(pct_developed)[, 1]
  )
```

```{r cv-function}
# Cross-validation helper: returns mean and SD of AUC across k folds
run_cv_model <- function(data, formula, k = 5) {
  folds <- rsample::vfold_cv(data, v = k, strata = is_present)

  results <- map_df(folds$splits, function(split) {
    train <- rsample::analysis(split)
    test  <- rsample::assessment(split)

    model <- glm(formula, data = train, family = binomial())
    probs <- predict(model, newdata = test, type = "response")
    auc_val <- pROC::auc(test$is_present, probs)

    tibble(auc = as.numeric(auc_val))
  })

  tibble(mean_auc = mean(results$auc), sd_auc = sd(results$auc))
}
```

```{r define-models}
model_list <- list(
  # Single-predictor models
  S1_pop       = is_present ~ z_pop,
  S2_radiance  = is_present ~ z_radiance,
  S3_dev       = is_present ~ z_dev,

  # Population density + one additional predictor
  M1_pop_rad   = is_present ~ z_pop + z_radiance,
  M2_pop_dev   = is_present ~ z_pop + z_dev,

  # Full urban footprint
  M4_pop_rad_dev = is_present ~ z_pop + z_radiance + z_dev
)
```

```{r run-cv, cache=TRUE}
species_results <- df_model %>%
  group_split(species) %>%
  map_df(function(sp_data) {
    sp_name <- unique(sp_data$species)
    map_df(names(model_list), function(mname) {
      cv_res <- run_cv_model(sp_data, model_list[[mname]])
      tibble(
        species  = sp_name,
        model    = mname,
        mean_auc = cv_res$mean_auc,
        sd_auc   = cv_res$sd_auc
      )
    })
  })
```

```{r auc-table}
# Key comparison: population density alone vs. additive models
species_results %>%
  filter(model %in% c("S1_pop", "M1_pop_rad", "M2_pop_dev", "M4_pop_rad_dev")) %>%
  mutate(
    model = recode(model,
      "S1_pop"         = "Pop. density only",
      "M1_pop_rad"     = "+ radiance",
      "M2_pop_dev"     = "+ development",
      "M4_pop_rad_dev" = "+ radiance + development"
    )
  ) %>%
  select(species, model, mean_auc, sd_auc) %>%
  arrange(species, desc(mean_auc))
```

```{r auc-barplot, fig.width=10, fig.height=5}
species_results %>%
  filter(model %in% c("S1_pop", "M1_pop_rad", "M2_pop_dev", "M4_pop_rad_dev")) %>%
  mutate(
    model = recode(model,
      "S1_pop"         = "Pop. density only",
      "M1_pop_rad"     = "+ nighttime light",
      "M2_pop_dev"     = "+ developed land",
      "M4_pop_rad_dev" = "+ light + development"
    ),
    model = factor(model, levels = c(
      "+ light + development",
      "+ developed land",
      "+ nighttime light",
      "Pop. density only"
    )),
    highlight = model == "Pop. density only"
  ) %>%
  ggplot(aes(x = mean_auc, y = model, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = sprintf("%.3f", mean_auc)),
    hjust = -0.15, size = 3.5
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#E63946", "FALSE" = "gray50"),
    guide = "none"
  ) +
  facet_wrap(~ species) +
  coord_cartesian(xlim = c(0.5, 1.0)) +
  labs(
    title = "Population density alone is nearly as predictive as multi-variable models",
    subtitle = "5-fold cross-validated AUC by species",
    x = "Mean AUC",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "italic", size = 11)
  )
```

### 4.2 Key finding: urban predictors are redundant at this scale

Population density alone achieves cross-validated AUC scores of
approximately 0.76–0.80 depending on species. Adding nighttime light
radiance, percent developed land, or both improves AUC by less than 1
percentage point in every case. At 10 km resolution, these variables are
effectively measuring the same underlying gradient of human modification
— they are interchangeable as predictors of bat presence.

This does not mean light and development are ecologically irrelevant. It
means that at this spatial scale, with opportunistic data, their effects
cannot be separated from the broader urban intensity signal. A
finer-resolution study with systematic survey data might be able to
tease them apart.

For the remaining analysis, we use **population density as the single
urban intensity predictor** — it is the most parsimonious, has the
clearest dual interpretation (accessibility proxy + human footprint),
and performs as well as any combination.

## 5. Do species respond differently to the urban gradient?

The EDA and model comparison established that population density
captures the dominant signal in these data. The next question is whether
species differ in *how strongly* they respond to that gradient.

### 5.1 Species × population density interaction model

```{r interaction-model}
# Factor species for interaction modeling
df_model <- df_model %>%
  mutate(species = factor(species))

# Fit interaction model: allows each species a different intercept and slope
model_interaction <- glm(
  is_present ~ species * z_pop,
  data = df_model,
  family = binomial()
)

summary(model_interaction)
```

### 5.2 Model comparison: does the interaction matter?

```{r model-comparison-aic}
model_pop         <- glm(is_present ~ z_pop, data = df_model, family = binomial())
model_pop_species <- glm(is_present ~ z_pop + species, data = df_model, family = binomial())

tibble(
  Model = c("Pop. density only",
            "Pop. + species (additive)",
            "Species \u00d7 pop. (interaction)"),
  AIC   = c(AIC(model_pop),
            AIC(model_pop_species),
            AIC(model_interaction)),
  df    = c(model_pop$df.residual,
            model_pop_species$df.residual,
            model_interaction$df.residual)
) %>%
  mutate(delta_AIC = AIC - min(AIC))
```

The interaction model provides the best fit (lowest AIC), confirming
that species differ not just in baseline presence rates but in the
strength of their association with population density.

### 5.3 Species-specific slopes and odds ratios

```{r odds-ratios}
coefs    <- coef(model_interaction)
vcov_mat <- vcov(model_interaction)

# Helper: extract species-specific slope and standard error
get_species_slope <- function(sp) {
  if (sp == "Aeorestes cinereus") {
    slope <- coefs["z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"]
  } else {
    int_term <- paste0("species", sp, ":z_pop")
    slope <- coefs["z_pop"] + coefs[int_term]
    var   <- vcov_mat["z_pop", "z_pop"] +
             vcov_mat[int_term, int_term] +
             2 * vcov_mat["z_pop", int_term]
  }
  tibble(species = sp, log_odds_slope = slope, se = sqrt(var))
}

or_results <- bind_rows(
  get_species_slope("Aeorestes cinereus"),
  get_species_slope("Myotis californicus"),
  get_species_slope("Myotis yumanensis")
) %>%
  mutate(
    odds_ratio = exp(log_odds_slope),
    lower_ci   = exp(log_odds_slope - 1.96 * se),
    upper_ci   = exp(log_odds_slope + 1.96 * se)
  )

or_results
```

```{r odds-ratio-plot, fig.width=7, fig.height=4}
ggplot(or_results, aes(x = odds_ratio, y = species)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0.2) +
  labs(
    title = "All three species show positive urban associations, but magnitudes differ",
    subtitle = "Odds ratio per 1 SD increase in log(population density), 95% CI",
    x = "Odds ratio",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.y = element_text(face = "italic"))
```

All three species have odds ratios significantly greater than 1, meaning
the probability of observing any of them increases with population
density. However, the magnitude of the effect differs: *Aeorestes
cinereus* and *Myotis yumanensis* show the strongest associations, while
*Myotis californicus* shows a notably weaker response to the urban
gradient.

### 5.4 Predicted probability curves

The most interpretable summary of the interaction model is a plot of
predicted presence probability across the population density gradient,
with the x-axis back-transformed to real-world units.

```{r predicted-curves, fig.width=9, fig.height=6}
# Back-transformation from z-score to population density
mu  <- mean(df_model$log1p_pop_density)
sig <- sd(df_model$log1p_pop_density)
z_to_pop <- function(z) exp((z * sig) + mu) - 1

# Build prediction grid
pred_grid <- expand.grid(
  z_pop   = seq(min(df_model$z_pop), max(df_model$z_pop), length.out = 200),
  species = levels(df_model$species)
)
pred_grid$pred_prob <- predict(
  model_interaction, newdata = pred_grid, type = "response"
)

# Reorder species for legend clarity
pred_grid$species <- factor(pred_grid$species, levels = c(
  "Aeorestes cinereus",
  "Myotis yumanensis",
  "Myotis californicus"
))

ggplot(pred_grid, aes(z_pop, pred_prob, color = species)) +
  geom_line(linewidth = 2) +
  scale_color_manual(values = species_colors) +
  scale_x_continuous(
    name = "Population density (people per km\u00B2, log-scaled)",
    labels = function(x) format(round(z_to_pop(x)), big.mark = ",")
  ) +
  labs(
    title = "Species differ in their association with the human-modified landscape gradient",
    subtitle = "Predicted probability of presence from species \u00d7 population density interaction model",
    y = "Predicted probability of presence",
    color = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.line = element_line(linewidth = 0.8, color = "black"),
    panel.grid.major = element_line(linewidth = 0.5, color = "gray85"),
    panel.grid.minor = element_blank(),
    legend.position = c(0.25, 0.80),
    legend.background = element_rect(fill = "white", color = "gray85"),
    legend.text = element_text(face = "italic")
  )
```

This figure — the central result of the analysis — shows that:

-   **All three species** have higher predicted presence probability in
    more urbanized areas, consistent with observer bias driving the
    overall pattern.
-   ***Aeorestes cinereus*** shows the highest predicted probabilities
    across the gradient, reaching approximately 30% in the most densely
    populated cells.
-   ***Myotis californicus*** shows the weakest response — its predicted
    probability curve rises more slowly and remains lower than the other
    two species even at high population densities.
-   The **divergence between species** is the ecologically interesting
    signal. Even though observer bias inflates all three curves, the
    *relative differences* between species are informative: they suggest
    that *Myotis californicus* may be more sensitive to urbanization, or
    at least less associated with human-modified landscapes, than the
    other two species.

## 6. Discussion

### 6.1 Summary of findings

This analysis used 455 GBIF bat occurrence records across three focal
species, mapped onto a 10 km equal-area grid covering accessible regions
of California, to examine how components of the human-modified landscape
structure bat observation patterns. Three main findings emerged:

1.  **Population density alone is sufficient to predict bat presence at
    this scale.** Nighttime light radiance and percent developed land
    are strongly collinear with population density at 10 km resolution
    and add negligible predictive value (less than 1% AUC improvement).
    At this scale, these variables function as interchangeable urban
    intensity proxies.

2.  **All three species show positive associations with increasing urban
    intensity.** This is expected given the opportunistic nature of GBIF
    data — more people means more observers means more bat records,
    regardless of true ecological preference.

3.  **Species differ in the strength of their urban association.**
    *Myotis californicus* shows a significantly weaker association with
    population density than either *Aeorestes cinereus* or *Myotis
    yumanensis*. This species-level difference persists even after
    accounting for the shared effect of observer bias, suggesting a
    genuine ecological signal: *Myotis californicus* may be more
    sensitive to urbanization or less tolerant of human-modified
    landscapes.

### 6.2 The dual role of population density

Population density plays two roles in this analysis, and the tension
between them is worth making explicit. First, it serves as an
**accessibility proxy**: areas with more people have more observers, so
GBIF records are biased toward populated areas. Second, it serves as a
**human footprint measure**: population density genuinely correlates
with habitat modification, light pollution, noise, and other stressors
that affect bat ecology.

These roles cannot be fully separated with opportunistic data. The fact
that all three species show positive associations with population
density likely reflects both accessibility bias and some degree of
genuine urban tolerance. The more informative result is the *difference*
between species: if all three were equally biased by observer effort,
their curves would be parallel. The fact that *Myotis californicus*
diverges suggests something ecologically real.

### 6.3 Limitations and future directions

-   **Observer bias remains the dominant confound.** Without systematic
    survey data (e.g., acoustic monitoring arrays with standardized
    effort), it is not possible to determine whether species are truly
    more common near urban areas or simply more frequently reported
    there.

-   **Spatial autocorrelation is not modeled.** Nearby grid cells are
    likely more similar than distant ones, violating the independence
    assumption of standard logistic regression. A spatial model (e.g.,
    using INLA or spatial random effects) could account for this.

-   **Temporal dynamics are collapsed.** Aggregating across 2012–2024
    treats presence as static, which obscures potential range shifts or
    changes in observation effort over time.

-   **Sample size is modest.** With 455 total records (and only 114 for
    *Myotis californicus*), effect estimates carry substantial
    uncertainty. Additional data sources (e.g., iNaturalist, state
    wildlife agency databases) could improve statistical power.

Despite these limitations, the analysis demonstrates a reproducible
framework for using open biodiversity data to detect species-specific
responses to landscape modification — an approach that scales to other
taxa and regions.

## 7. Data sources and citations

| Dataset | Source | Access |
|----|----|----|
| Bat occurrence records | GBIF (2012–2024) | [gbif.org](https://www.gbif.org/) |
| Nighttime light radiance | VIIRS VNL v2.1/v2.2, EOG Payne Institute | [eogdata.mines.edu](https://eogdata.mines.edu/products/vnl/) |
| Land cover | NLCD 2019 | [mrlc.gov](https://www.mrlc.gov/) |
| Protected areas | PAD-US 4.1 | [usgs.gov/programs/gap-analysis-project](https://www.usgs.gov/programs/gap-analysis-project/science/pad-us-data-download) |
| Population density | GPWv4 rev. 11 (SEDAC/CIESIN) 2020 | [earthdata.nasa.gov](https://www.earthdata.nasa.gov/data/catalog/sedac-ciesin-sedac-gpwv4-popdens-r11-4.11) |
| State boundary | US Census TIGER/Line | via `tigris` R package |

```{r session-info}
sessionInfo()
```
