---
title: "How Do Human-Modified Landscapes Shape Bat Observations in California?"
subtitle: "Environmental Data Science Capstone Project"
author: "Michele Perry"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: cosmo
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center"
)

options(scipen = 999, dplyr.summarise.inform = FALSE)
set.seed(123)
```

## 1. Introduction

Urbanization reshapes landscapes at every scale — replacing habitat with
impervious surface, flooding the night sky with artificial light, and
concentrating human activity into dense corridors. For wildlife, these
changes create winners and losers. Some species thrive at the urban
edge; others retreat to darker, less disturbed landscapes. Understanding
which species fall into which category is central to managing
biodiversity in an increasingly human-dominated world.

Bats are a particularly informative group for studying these dynamics.
As nocturnal insectivores, they are directly affected by artificial
light (which can disrupt foraging behavior and prey availability), land
development (which removes roosting habitat), and the broader footprint
of urbanization. California, with its steep urban-to-wildland gradients
and rich bat fauna, provides an ideal setting to ask how these pressures
structure species distributions.

This project examines three focal bat species with contrasting
ecological profiles:

-   *Aeorestes cinereus* (hoary bat) — a long-distance migratory,
    tree-roosting species
-   *Myotis yumanensis* (Yuma myotis) — a water-associated species often
    found near bridges and buildings
-   *Myotis californicus* (California myotis) — a small,
    crevice-roosting species widespread across the state

Using publicly available occurrence records from the [Global
Biodiversity Information Facility (GBIF)](https://www.gbif.org/), this
study asks:

> **How do components of human-modified landscapes structure bat
> observation patterns at fine spatial scales, and do species respond
> differently to increasing urban intensity?**

### A critical framing note: observer bias in opportunistic data

GBIF records are opportunistic — they reflect where people *looked*, not
just where bats *are*. Observations cluster near roads, cities, and
research stations, meaning that any apparent association between bats
and urbanization could partly reflect reporting effort rather than true
ecological preference. This project addresses this bias explicitly by
including population density as an accessibility proxy and by
interpreting results as patterns in *observed* presence rather than
confirmed ecological effects.

## 2. Data acquisition and processing

The data processing pipeline lives in the `R/` directory as a series of
modular scripts (`R/00_setup.R` through
`R/06_build_grid_model_dataset.R`) that can be executed in sequence to
reproduce the full workflow from raw data to analysis-ready dataset.
This section walks through the key decisions and transformations rather
than re-running the pipeline; readers interested in implementation
details should consult the scripts directly.

### 2.1 GBIF bat occurrence data

Occurrence records for the three focal species were downloaded from GBIF
for the period 2012–2024, filtered to records with valid coordinates.
The temporal scope aligns with the availability of consistently
processed VIIRS nighttime lights satellite products (2012–present).

Key cleaning steps (implemented in `R/01_get_clean_bat_points.R`):

-   Selected a minimal set of fields for analysis and provenance
    tracking
-   Removed records with missing, zero, or out-of-range coordinates
-   Applied a broad California bounding box as an initial spatial
    guardrail
-   Resolved a taxonomic synonymy: GBIF normalizes *Lasiurus cinereus*
    to *Aeorestes cinereus* in its backbone taxonomy, so filtering was
    performed on `species_key` rather than name strings to ensure
    consistency
-   Converted records to spatial points (WGS 84) for downstream spatial
    operations

The cleaned points were then spatially filtered to the California state
boundary (`R/03_build_grid_and_domain.R`), producing 455 unique
occurrence records across the three species.

```{r load-libraries}
library(tidyverse)
library(sf)
library(scales)
library(viridis)
library(here)
library(broom)
library(pROC)
library(rsample)
library(purrr)
```

```{r load-data}
# Final analysis-ready dataset (cell_id x species panel with covariates)
df <- read_csv(
  here("data", "processed", "analysis_grid", "grid_model_dataset_10km.csv"),
  show_col_types = FALSE
)

# Spatial files for mapping
grid <- st_read(
  here("data", "processed", "grid", "ca_grid10km_accessible.gpkg"),
  quiet = TRUE
)
pts <- st_read(
  here("data", "processed", "gbif", "gbif_bats_points_clean_CA_2012_2024.gpkg"),
  quiet = TRUE
)
ca <- st_read(
  here("data", "processed", "boundaries", "ca_boundary.gpkg"),
  layer = "ca_boundary",
  quiet = TRUE
)
```

Let's verify the structure of the final dataset:

```{r inspect-data}
glimpse(df)
```

```{r data-summary}
# Quick summary: how many cells, species, and observations?
df %>%
  summarise(
    n_cells = n_distinct(cell_id),
    n_species = n_distinct(species),
    panel_rows = n(),
    total_observations = sum(n_obs),
    overall_presence_rate = mean(is_present)
  )
```

```{r species-summary}
# Per-species breakdown
df %>%
  group_by(species) %>%
  summarise(
    n_cells = n(),
    n_present = sum(is_present),
    share_present = round(mean(is_present), 4),
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_present))
```

The dataset is a balanced panel of 2,811 grid cells × 3 species = 8,433
rows. Each row records whether any observations of a given species
occurred in that cell across the full 2012–2024 study period, along with
four landscape covariates.

### 2.2 Spatial framework: 10 km equal-area grid

Rather than using administrative boundaries (e.g., counties), this
analysis uses a 10 km equal-area grid in California Albers projection
(EPSG:3310). A regular grid avoids conflating administrative boundaries
with ecological gradients — counties vary enormously in area and span
multiple habitat types, which can mask fine-scale landscape
associations.

The 10 km resolution was chosen as a practical balance: fine enough to
capture meaningful variation in landscape structure within a single
county, but coarse enough that each cell contains sufficient area for
the landscape covariates (nighttime light, land cover, population
density) to be meaningfully summarized.

**Accessible modeling domain.** Not all of California has been surveyed
for bats. To avoid treating unsampled areas as ecological absences, the
analysis is restricted to an "accessible domain" defined by buffering
all bat observation locations by 50 km and taking the union. Grid cells
with centroids inside this domain (plus any cell containing an
observation) form the final set of 2,811 accessible cells — roughly 75%
of the full California grid.

### 2.3 Landscape covariates

Four covariates were computed at the grid-cell level (implemented in
`R/04_build_grid_covariates.R`):

| Variable | Source | Description |
|-----------------------|-------------------|------------------------------|
| `mean_radiance` | VIIRS VNL v2.1/v2.2 | Mean nighttime light radiance, averaged across annual composites (2012–2024) |
| `pct_developed` | NLCD 2019 | Proportion of cell classified as developed land (all "Developed" classes) |
| `pct_protected` | PAD-US 4.1 | Proportion of cell under GAP Status 1–3 protection |
| `pop_density` | WorldPop 2020 | Mean population density (people per km²) |

Log-transformed versions (`log1p_radiance`, `log1p_pop_density`) are
used throughout the analysis to reduce right skew in the urban-dominated
variables.

**Why these four?** Nighttime light, land development, and population
density each capture a different facet of the human-modified landscape.
Protected land coverage provides a counterpoint — areas where the
landscape is relatively intact. Together, they characterize a gradient
from wild, dark, unpopulated landscapes to bright, developed, densely
inhabited ones.

### 2.4 Presence aggregation

Because the environmental covariates show far stronger spatial than
temporal variation (nighttime light patterns are largely stable
year-to-year at this scale), bat records were aggregated across the full
study period into cumulative spatial presence: each cell × species
combination is coded as 1 (at least one observation in 2012–2024) or 0
(none). This collapses a sparse, zero-heavy count distribution into a
cleaner binary response suitable for logistic regression.

## 3. Exploratory data analysis

### 3.1 Response structure: extreme sparsity

```{r presence-absence-bar}
df %>%
  mutate(
    status = if_else(is_present == 1, "Present (\u22651 record)", "Absent (0 records)")
  ) %>%
  count(species, status) %>%
  group_by(species) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = species, y = prop, fill = status)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("Present (\u22651 record)" = "#1D9E75",
               "Absent (0 records)" = "gray75")
  ) +
  labs(
    title = "Most grid cells have no bat observations",
    subtitle = "Proportion of 2,811 accessible cells with at least one record (2012\u20132024)",
    x = NULL,
    y = "Proportion of cells",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

All three species are observed in fewer than 5% of grid cells.
*Aeorestes cinereus* is the most frequently observed (4.6% of cells),
while *Myotis californicus* is the rarest (2.4%). This extreme sparsity
reinforces the decision to model binary presence rather than observation
counts — any count-based model would be overwhelmed by structural zeros.

### 3.2 Mapping observations over the human-modified landscape

```{r prepare-map-data}
# Join covariates to the spatial grid for mapping
grid_map <- grid %>%
  mutate(cell_id = as.character(cell_id)) %>%
  left_join(
    df %>%
      mutate(cell_id = as.character(cell_id)) %>%
      distinct(cell_id, log1p_pop_density, log1p_radiance, pct_developed),
    by = "cell_id"
  )

# Align CRS for all layers
pts <- st_transform(pts, st_crs(grid_map))
ca  <- st_transform(ca, st_crs(grid_map))
```

```{r map-pop-density, fig.width=8, fig.height=9}
ggplot() +
  geom_sf(data = grid_map, aes(fill = log1p_pop_density), color = NA) +
  scale_fill_viridis_c(
    option = "cividis",
    name = "Population density\n(people/km\u00B2, log-scaled)",
    breaks = c(0, 2, 4, 6, 8),
    labels = c("0", "10", "50", "400", "3,000")
  ) +
  geom_sf(data = ca, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts, color = "#2CBEC9", size = 1.5, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "GBIF bat observations cluster near urban areas",
    subtitle = "Cyan points = bat observations; background = population density on accessible 10 km grid"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.8, "cm")
  )
```

The map reveals a clear pattern: bat observations are concentrated in
and around California's major population centers — the San Francisco Bay
Area, Los Angeles basin, San Diego, and Sacramento. Large swaths of the
state's interior, mountains, and deserts have few or no records. This is
the observer bias problem in action: people (and therefore observation
effort) concentrate in cities.

```{r map-radiance, fig.width=8, fig.height=9}
ggplot() +
  geom_sf(data = grid_map, aes(fill = log1p_radiance), color = NA) +
  scale_fill_viridis_c(
    option = "cividis",
    name = "Nighttime light\n(log radiance)"
  ) +
  geom_sf(data = ca, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts, color = "#2CBEC9", size = 1.5, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "Bat observations over nighttime light intensity",
    subtitle = "VIIRS radiance averaged across 2012\u20132024, aggregated to 10 km grid"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.8, "cm")
  )
```

```{r map-developed, fig.width=8, fig.height=9}
ggplot() +
  geom_sf(data = grid_map, aes(fill = pct_developed), color = NA) +
  scale_fill_viridis_c(
    option = "cividis",
    name = "% developed land",
    labels = percent_format(accuracy = 1)
  ) +
  geom_sf(data = ca, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts, color = "#2CBEC9", size = 1.5, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "Bat observations over developed land cover",
    subtitle = "NLCD 2019 developed classes, aggregated to 10 km grid"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.8, "cm")
  )
```

All three maps tell a strikingly similar story: the spatial patterns of
nighttime light, land development, and population density largely
overlap. Bat observations track all three. This visual impression of
collinearity among the urban predictors will become important in the
modeling section.

### 3.3 Predictor correlations

```{r correlation-matrix}
cov_df <- df %>%
  select(log1p_radiance, log1p_pop_density, pct_developed, pct_protected) %>%
  drop_na()

cor_mat <- round(cor(cov_df), 3)
cor_mat
```

```{r predictor-distributions, fig.height=5}
df %>%
  select(log1p_radiance, log1p_pop_density, pct_developed, pct_protected) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "gray50", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Distributions of landscape predictors",
    subtitle = "Most cells are dark, undeveloped, and sparsely populated",
    x = NULL, y = "Count"
  )
```

Nighttime light, population density, and percent developed land are all
strongly correlated (r = 0.76–0.87). At 10 km resolution, these
variables essentially capture the same underlying gradient of human
modification. Percent protected land is moderately negatively correlated
with the urban variables, as expected — protected areas tend to be less
developed — but the relationship is weaker.

This collinearity has an important analytical implication: including all
three urban proxies in a single model would introduce redundancy without
adding predictive value.

### 3.4 Presence rates along landscape gradients

To explore how bat presence varies with each predictor, we divide each
variable into decile bins and compute the share of cells with at least
one observation in each bin, separately by species.

```{r species-colors}
# Consistent species color palette used throughout
species_colors <- c(
  "Aeorestes cinereus"  = "#1B9E77",
  "Myotis yumanensis"   = "#D95F02",
  "Myotis californicus" = "#7570B3"
)
```

```{r presence-vs-pop, fig.height=5}
df %>%
  mutate(bin = ntile(log1p_pop_density, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_pop_density, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = species_colors) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Presence rate increases with population density",
    subtitle = "Binned deciles of log(1 + population density)",
    x = "Mean log(1 + population density) within bin",
    y = "Presence rate",
    color = "Species"
  )
```

```{r presence-vs-radiance, fig.height=5}
df %>%
  mutate(bin = ntile(log1p_radiance, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_radiance, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = species_colors) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Presence rate increases with nighttime light intensity",
    subtitle = "Binned deciles of log(1 + radiance)",
    x = "Mean log(1 + radiance) within bin",
    y = "Presence rate",
    color = "Species"
  )
```

```{r presence-vs-developed, fig.height=5}
df %>%
  mutate(bin = ntile(pct_developed, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(pct_developed, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = species_colors) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Presence rate increases with land development",
    subtitle = "Binned deciles of percent developed land",
    x = "Mean percent developed within bin",
    y = "Presence rate",
    color = "Species"
  )
```

All three urban-associated predictors show the same qualitative pattern:
presence rates rise sharply in the upper deciles, and species differ in
the strength of that rise. *Aeorestes cinereus* consistently shows the
steepest increase, while *Myotis californicus* shows a shallower
response. These exploratory patterns motivate the formal modeling that
follows.

## 4. Model comparison: what predicts bat presence?

Given the strong collinearity among urban predictors, a natural question
is: **does adding nighttime light or land development improve
predictions beyond population density alone?** This matters both
practically (simpler models are easier to interpret) and conceptually
(if the variables are interchangeable, they're measuring the same
thing).

### 4.1 Cross-validated model comparison

We fit a series of logistic regression models predicting binary presence
(`is_present`) and compare their performance using 5-fold
cross-validated AUC (area under the ROC curve). Models are fit
separately for each species to allow species-specific effects.

```{r standardize-predictors}
# Standardize predictors for comparable coefficients
df_model <- df %>%
  mutate(
    z_radiance  = scale(log1p_radiance)[, 1],
    z_pop       = scale(log1p_pop_density)[, 1],
    z_dev       = scale(pct_developed)[, 1],
    z_protected = scale(pct_protected)[, 1]
  )
```

```{r cv-function}
# Cross-validation helper: returns mean and SD of AUC across k folds
run_cv_model <- function(data, formula, k = 5) {
  folds <- rsample::vfold_cv(data, v = k, strata = is_present)

  results <- map_df(folds$splits, function(split) {
    train <- rsample::analysis(split)
    test  <- rsample::assessment(split)

    model <- glm(formula, data = train, family = binomial())
    probs <- predict(model, newdata = test, type = "response")
    auc_val <- pROC::auc(test$is_present, probs)

    tibble(auc = as.numeric(auc_val))
  })

  tibble(mean_auc = mean(results$auc), sd_auc = sd(results$auc))
}
```

```{r define-models}
model_list <- list(
  # Single-predictor models
  S1_pop       = is_present ~ z_pop,
  S2_radiance  = is_present ~ z_radiance,
  S3_dev       = is_present ~ z_dev,
  S4_protected = is_present ~ z_protected,

  # Population density + one additional predictor
  M1_pop_rad   = is_present ~ z_pop + z_radiance,
  M2_pop_dev   = is_present ~ z_pop + z_dev,
  M3_pop_prot  = is_present ~ z_pop + z_protected,

  # Full urban footprint
  M4_pop_rad_dev      = is_present ~ z_pop + z_radiance + z_dev,
  M5_full_no_interact = is_present ~ z_pop + z_radiance + z_dev + z_protected
)
```

```{r run-cv, cache=TRUE}
species_results <- df_model %>%
  group_split(species) %>%
  map_df(function(sp_data) {
    sp_name <- unique(sp_data$species)
    map_df(names(model_list), function(mname) {
      cv_res <- run_cv_model(sp_data, model_list[[mname]])
      tibble(
        species  = sp_name,
        model    = mname,
        mean_auc = cv_res$mean_auc,
        sd_auc   = cv_res$sd_auc
      )
    })
  })
```

```{r auc-table}
# Key comparison: population density alone vs. additive models
species_results %>%
  filter(model %in% c("S1_pop", "M1_pop_rad", "M2_pop_dev", "M4_pop_rad_dev")) %>%
  mutate(
    model = recode(model,
      "S1_pop"         = "Pop. density only",
      "M1_pop_rad"     = "+ radiance",
      "M2_pop_dev"     = "+ development",
      "M4_pop_rad_dev" = "+ radiance + development"
    )
  ) %>%
  select(species, model, mean_auc, sd_auc) %>%
  arrange(species, desc(mean_auc))
```

```{r auc-barplot, fig.width=10, fig.height=5}
species_results %>%
  filter(model %in% c("S1_pop", "M1_pop_rad", "M2_pop_dev", "M4_pop_rad_dev")) %>%
  mutate(
    model = recode(model,
      "S1_pop"         = "Pop. density only",
      "M1_pop_rad"     = "+ nighttime light",
      "M2_pop_dev"     = "+ developed land",
      "M4_pop_rad_dev" = "+ light + development"
    ),
    model = factor(model, levels = c(
      "+ light + development",
      "+ developed land",
      "+ nighttime light",
      "Pop. density only"
    )),
    highlight = model == "Pop. density only"
  ) %>%
  ggplot(aes(x = mean_auc, y = model, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = sprintf("%.3f", mean_auc)),
    hjust = -0.15, size = 3.5
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#E63946", "FALSE" = "gray50"),
    guide = "none"
  ) +
  facet_wrap(~ species) +
  coord_cartesian(xlim = c(0.5, 1.0)) +
  labs(
    title = "Population density alone is nearly as predictive as multi-variable models",
    subtitle = "5-fold cross-validated AUC by species",
    x = "Mean AUC",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "italic", size = 11)
  )
```

### 4.2 Key finding: urban predictors are redundant at this scale

Population density alone achieves cross-validated AUC scores of
approximately 0.76–0.80 depending on species. Adding nighttime light
radiance, percent developed land, or both improves AUC by less than 1
percentage point in every case. At 10 km resolution, these variables are
effectively measuring the same underlying gradient of human modification
— they are interchangeable as predictors of bat presence.

This does not mean light and development are ecologically irrelevant. It
means that at this spatial scale, with opportunistic data, their effects
cannot be separated from the broader urban intensity signal. A
finer-resolution study with systematic survey data might be able to
tease them apart.

For the remaining analysis, we use **population density as the single
urban intensity predictor** — it is the most parsimonious, has the
clearest dual interpretation (accessibility proxy + human footprint),
and performs as well as any combination.

## 5. Do species respond differently to the urban gradient?

The EDA and model comparison established that population density
captures the dominant signal in these data. The next question is whether
species differ in *how strongly* they respond to that gradient.

### 5.1 Species × population density interaction model

```{r interaction-model}
# Factor species for interaction modeling
df_model <- df_model %>%
  mutate(species = factor(species))

# Fit interaction model: allows each species a different intercept and slope
model_interaction <- glm(
  is_present ~ species * z_pop,
  data = df_model,
  family = binomial()
)

summary(model_interaction)
```

### 5.2 Model comparison: does the interaction matter?

```{r model-comparison-aic}
model_pop         <- glm(is_present ~ z_pop, data = df_model, family = binomial())
model_pop_species <- glm(is_present ~ z_pop + species, data = df_model, family = binomial())

tibble(
  Model = c("Pop. density only",
            "Pop. + species (additive)",
            "Species \u00d7 pop. (interaction)"),
  AIC   = c(AIC(model_pop),
            AIC(model_pop_species),
            AIC(model_interaction)),
  df    = c(model_pop$df.residual,
            model_pop_species$df.residual,
            model_interaction$df.residual)
) %>%
  mutate(delta_AIC = AIC - min(AIC))
```

The interaction model provides the best fit (lowest AIC), confirming
that species differ not just in baseline presence rates but in the
strength of their association with population density.

### 5.3 Species-specific slopes and odds ratios

```{r odds-ratios}
coefs    <- coef(model_interaction)
vcov_mat <- vcov(model_interaction)

# Helper: extract species-specific slope and standard error
get_species_slope <- function(sp) {
  if (sp == "Aeorestes cinereus") {
    slope <- coefs["z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"]
  } else {
    int_term <- paste0("species", sp, ":z_pop")
    slope <- coefs["z_pop"] + coefs[int_term]
    var   <- vcov_mat["z_pop", "z_pop"] +
             vcov_mat[int_term, int_term] +
             2 * vcov_mat["z_pop", int_term]
  }
  tibble(species = sp, log_odds_slope = slope, se = sqrt(var))
}

or_results <- bind_rows(
  get_species_slope("Aeorestes cinereus"),
  get_species_slope("Myotis californicus"),
  get_species_slope("Myotis yumanensis")
) %>%
  mutate(
    odds_ratio = exp(log_odds_slope),
    lower_ci   = exp(log_odds_slope - 1.96 * se),
    upper_ci   = exp(log_odds_slope + 1.96 * se)
  )

or_results
```

```{r odds-ratio-plot, fig.width=7, fig.height=4}
ggplot(or_results, aes(x = odds_ratio, y = species)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0.2) +
  labs(
    title = "All three species show positive urban associations, but magnitudes differ",
    subtitle = "Odds ratio per 1 SD increase in log(population density), 95% CI",
    x = "Odds ratio",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.y = element_text(face = "italic"))
```

All three species have odds ratios significantly greater than 1, meaning
the probability of observing any of them increases with population
density. However, the magnitude of the effect differs: *Aeorestes
cinereus* and *Myotis yumanensis* show the strongest associations, while
*Myotis californicus* shows a notably weaker response to the urban
gradient.

### 5.4 Predicted probability curves

The most interpretable summary of the interaction model is a plot of
predicted presence probability across the population density gradient,
with the x-axis back-transformed to real-world units.

```{r predicted-curves, fig.width=9, fig.height=6}
# Back-transformation from z-score to population density
mu  <- mean(df_model$log1p_pop_density)
sig <- sd(df_model$log1p_pop_density)
z_to_pop <- function(z) exp((z * sig) + mu) - 1

# Build prediction grid
pred_grid <- expand.grid(
  z_pop   = seq(min(df_model$z_pop), max(df_model$z_pop), length.out = 200),
  species = levels(df_model$species)
)
pred_grid$pred_prob <- predict(
  model_interaction, newdata = pred_grid, type = "response"
)

# Reorder species for legend clarity
pred_grid$species <- factor(pred_grid$species, levels = c(
  "Aeorestes cinereus",
  "Myotis yumanensis",
  "Myotis californicus"
))

ggplot(pred_grid, aes(z_pop, pred_prob, color = species)) +
  geom_line(linewidth = 2) +
  scale_color_manual(values = species_colors) +
  scale_x_continuous(
    name = "Population density (people per km\u00B2, log-scaled)",
    labels = function(x) format(round(z_to_pop(x)), big.mark = ",")
  ) +
  labs(
    title = "Species differ in their association with the human-modified landscape gradient",
    subtitle = "Predicted probability of presence from species \u00d7 population density interaction model",
    y = "Predicted probability of presence",
    color = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.line = element_line(linewidth = 0.8, color = "black"),
    panel.grid.major = element_line(linewidth = 0.5, color = "gray85"),
    panel.grid.minor = element_blank(),
    legend.position = c(0.25, 0.80),
    legend.background = element_rect(fill = "white", color = "gray85"),
    legend.text = element_text(face = "italic")
  )
```

This figure — the central result of the analysis — shows that:

-   **All three species** have higher predicted presence probability in
    more urbanized areas, consistent with observer bias driving the
    overall pattern.
-   ***Aeorestes cinereus*** shows the highest predicted probabilities
    across the gradient, reaching approximately 30% in the most densely
    populated cells.
-   ***Myotis californicus*** shows the weakest response — its predicted
    probability curve rises more slowly and remains lower than the other
    two species even at high population densities.
-   The **divergence between species** is the ecologically interesting
    signal. Even though observer bias inflates all three curves, the
    *relative differences* between species are informative: they suggest
    that *Myotis californicus* may be more sensitive to urbanization, or
    at least less associated with human-modified landscapes, than the
    other two species.

## 6. Discussion

### 6.1 Summary of findings

This analysis used 455 GBIF bat occurrence records across three focal
species, mapped onto a 10 km equal-area grid covering accessible regions
of California, to examine how components of the human-modified landscape
structure bat observation patterns. Three main findings emerged:

1.  **Population density alone is sufficient to predict bat presence at
    this scale.** Nighttime light radiance and percent developed land
    are strongly collinear with population density at 10 km resolution
    and add negligible predictive value (less than 1% AUC improvement).
    At this scale, these variables function as interchangeable urban
    intensity proxies.

2.  **All three species show positive associations with increasing urban
    intensity.** This is expected given the opportunistic nature of GBIF
    data — more people means more observers means more bat records,
    regardless of true ecological preference.

3.  **Species differ in the strength of their urban association.**
    *Myotis californicus* shows a significantly weaker association with
    population density than either *Aeorestes cinereus* or *Myotis
    yumanensis*. This species-level difference persists even after
    accounting for the shared effect of observer bias, suggesting a
    genuine ecological signal: *Myotis californicus* may be more
    sensitive to urbanization or less tolerant of human-modified
    landscapes.

### 6.2 The dual role of population density

Population density plays two roles in this analysis, and the tension
between them is worth making explicit. First, it serves as an
**accessibility proxy**: areas with more people have more observers, so
GBIF records are biased toward populated areas. Second, it serves as a
**human footprint measure**: population density genuinely correlates
with habitat modification, light pollution, noise, and other stressors
that affect bat ecology.

These roles cannot be fully separated with opportunistic data. The fact
that all three species show positive associations with population
density likely reflects both accessibility bias and some degree of
genuine urban tolerance. The more informative result is the *difference*
between species: if all three were equally biased by observer effort,
their curves would be parallel. The fact that *Myotis californicus*
diverges suggests something ecologically real.

### 6.3 Limitations and future directions

-   **Observer bias remains the dominant confound.** Without systematic
    survey data (e.g., acoustic monitoring arrays with standardized
    effort), it is not possible to determine whether species are truly
    more common near urban areas or simply more frequently reported
    there.

-   **Spatial autocorrelation is not modeled.** Nearby grid cells are
    likely more similar than distant ones, violating the independence
    assumption of standard logistic regression. A spatial model (e.g.,
    using INLA or spatial random effects) could account for this.

-   **Temporal dynamics are collapsed.** Aggregating across 2012–2024
    treats presence as static, which obscures potential range shifts or
    changes in observation effort over time.

-   **Sample size is modest.** With 455 total records (and only 114 for
    *Myotis californicus*), effect estimates carry substantial
    uncertainty. Additional data sources (e.g., iNaturalist, state
    wildlife agency databases) could improve statistical power.

Despite these limitations, the analysis demonstrates a reproducible
framework for using open biodiversity data to detect species-specific
responses to landscape modification — an approach that scales to other
taxa and regions.

## 7. Data sources and citations

| Dataset | Source | Access |
|--------------------------|-----------------------|-----------------------|
| Bat occurrence records | GBIF (2012–2024) | [gbif.org](https://www.gbif.org/) |
| Nighttime light radiance | VIIRS VNL v2.1/v2.2, EOG Payne Institute | [eogdata.mines.edu](https://eogdata.mines.edu/products/vnl/) |
| Land cover | NLCD 2019 | [mrlc.gov](https://www.mrlc.gov/) |
| Protected areas | PAD-US 4.1 | [usgs.gov/programs/gap-analysis-project](https://www.usgs.gov/programs/gap-analysis-project/science/pad-us-data-download) |
| Population density | WorldPop 2020 | [worldpop.org](https://www.worldpop.org/) |
| State boundary | US Census TIGER/Line | via `tigris` R package |

```{r session-info}
sessionInfo()
```
