---
title: "How Do Human-Modified Landscapes Shape Bat Observations in California?"
subtitle: "Environmental Data Science Capstone Project"
author: "Michele Perry"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: cosmo
  github_document:
    toc: true
    html_preview: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center"
)

options(scipen = 999, dplyr.summarise.inform = FALSE)
set.seed(123)
```

```{css figure-spacing, echo=FALSE}
img {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}
```

## 1. Introduction

Urbanization reshapes landscapes at every scale — replacing habitat with
impervious surface, flooding the night sky with artificial light, and
concentrating human activity into dense corridors. For wildlife, these
changes create winners and losers. Some species thrive at the urban
edge; others retreat to darker, less disturbed landscapes. Understanding
which species fall into which category is central to managing
biodiversity in an increasingly human-dominated world.

Bats are a particularly informative group for studying these dynamics.
As nocturnal insectivores, they are directly affected by artificial
light (which can disrupt foraging behavior and prey availability), land
development (which removes roosting habitat), and the broader footprint
of urbanization. California, with its steep urban-to-wildland gradients
and rich bat fauna, provides an ideal setting to ask how these pressures
structure species distributions.

This project examines three focal bat species with contrasting
ecological profiles:

-   *Aeorestes cinereus* (hoary bat) — a long-distance migratory,
    tree-roosting species
-   *Myotis yumanensis* (Yuma myotis) — a water-associated species often
    found near bridges and buildings
-   *Myotis californicus* (California myotis) — a small,
    crevice-roosting species widespread across the state

These species were chosen because their contrasting roosting and
foraging ecologies suggest they may respond differently to urbanization.

Using publicly available occurrence records from the [Global
Biodiversity Information Facility (GBIF)](https://www.gbif.org/), this
study asks:

> **How do human-modified landscapes structure bat observation patterns
> at fine spatial scales, and do species respond differently to
> increasing urban intensity?**

### A note on observer bias

GBIF records are opportunistic — they reflect where people *looked*,
not just where bats *are*. Because observations cluster near roads,
cities, and research stations, apparent associations between bats and
urbanization may partly reflect reporting effort. This analysis accounts
for that bias by including population density as an accessibility proxy
and by interpreting results as patterns in *observed* presence rather
than confirmed ecological preference.

## 2. Data acquisition and processing

The data processing pipeline lives in the `R/` directory as a series of
modular scripts (`R/00_setup.R` through
`R/06_build_grid_model_dataset.R`) that can be run in sequence to
reproduce the full workflow from raw downloads to the final
analysis-ready dataset. This section walks through each stage in the
order the data were actually acquired and processed, highlighting key
spatial operations and design decisions along the way.

```{r load-libraries}
library(tidyverse)
library(sf)
library(scales)
library(viridis)
library(here)
library(broom)
library(pROC)
library(rsample)
library(purrr)
library(showtext)

font_add_google("Lato", "lato")
showtext_auto()
```

### 2.1 GBIF bat occurrence data

The starting point for this project is publicly available species
occurrence data from the [Global Biodiversity Information Facility
(GBIF)](https://www.gbif.org/). GBIF aggregates biodiversity records
from museum collections, citizen science platforms, and research surveys
into a single downloadable format.

The `rgbif` package provides an R interface to the GBIF API. The
workflow in `R/01_get_clean_bat_points.R` begins by resolving each focal
species name against the GBIF backbone taxonomy using `name_backbone()`,
which returns a unique `species_key` — a stable numeric identifier that
avoids issues with taxonomic synonyms.

The download request is submitted via `occ_download()` with filters for
the three focal species keys, valid coordinates, and years 2012–2024:

```{r gbif-download-example, eval=FALSE}
# (display only — not executed; see R/01_get_clean_bat_points.R)
dl <- occ_download(
  pred_in("taxonKey", taxa_keys),
  pred("hasCoordinate", TRUE),
  pred_gte("year", 2012),
  pred_lte("year", 2024),
  format = "SIMPLE_CSV"
)
```

GBIF processes this request server-side and returns a zipped CSV with
dozens of columns. The cleaning step selects only the fields needed for
this analysis and applies coordinate hygiene filters:

```{r cleaning-example, eval=FALSE}
# (display only — not executed; see R/01_get_clean_bat_points.R)
dat <- raw %>%
  transmute(
    gbif_id, species, scientific_name,
    year, month, day,
    lon = decimal_longitude,
    lat = decimal_latitude,
    coordinate_uncertainty_m = coordinate_uncertainty_in_meters
  ) %>%
  filter(
    !is.na(lon), !is.na(lat),        # must have coordinates
    lon != 0, lat != 0,               # exclude zero-island artifacts
    lon >= -125, lon <= -113,          # rough CA bounding box
    lat >= 32, lat <= 42
  )
```

After filtering, the cleaned records are converted to an `sf` spatial
points object in WGS 84 (EPSG:4326) using `st_as_sf()` — this is the
starting point for all downstream spatial operations.

### 2.2 California state boundary

A California state boundary polygon, retrieved from the US Census
TIGER/Line database via the `tigris` package (`R/02_get_ca_boundary.R`),
is used for spatially filtering bat points and constructing the analysis
grid.

```{r load-boundary-and-points}
ca <- st_read(
  here("data", "processed", "boundaries", "ca_boundary.gpkg"),
  layer = "ca_boundary", quiet = TRUE
)

pts <- st_read(
  here("data", "processed", "gbif", "gbif_bats_points_clean_CA_2012_2024.gpkg"),
  quiet = TRUE
)
```

```{r points-summary}
cat("Total bat occurrence records in California:", nrow(pts), "\n")
pts %>% st_drop_geometry() %>% count(species, sort = TRUE)
```

These 455 records are the result of filtering the raw GBIF download to
valid California coordinates and then performing a spatial intersection
with the state boundary (`st_intersects()` in
`R/03_build_grid_and_domain.R`).

### 2.3 Spatial framework: 10 km equal-area grid

With the bat points and boundary in hand, the next decision is how to
structure the analysis spatially. Rather than using administrative units
like counties (which vary enormously in size and span multiple habitat
types), this project uses a regular 10 km equal-area grid. A uniform
grid ensures that every analysis unit covers the same area, making
landscape summaries directly comparable across cells.

All spatial operations use the California Albers projection (EPSG:3310),
an equal-area coordinate system measured in meters — essential for any
analysis where area calculations matter.

The grid is constructed in `R/03_build_grid_and_domain.R` using
`sf::st_make_grid()`:

```{r grid-construction-example, eval=FALSE}
# (display only — not executed; see R/03_build_grid_and_domain.R)
grid_full <- st_make_grid(
  ca_ae,                     # CA boundary in Albers projection
  cellsize = 10000,          # 10 km cells
  square = TRUE
) %>%
  st_as_sf() %>%
  mutate(cell_id = row_number())
```

This produces a grid of regular squares covering the full bounding box
of California. The analysis grid is then trimmed using two criteria: (1)
cells whose centroids fall inside California, and (2) a "presence
rescue" step that retains any cell containing at least one bat
observation, even if its centroid falls just outside the boundary. This
prevents edge-case data loss along the coastline and state borders.

**Defining the accessible domain.** Not all of California has been
surveyed for bats. To avoid treating genuinely unsampled areas as
ecological absences, the analysis is restricted to an "accessible"
subset of the grid. This domain is defined by buffering every bat
observation point by 50 km using `st_buffer()`, taking the union of all
buffers, and intersecting the result with a buffered California
boundary. Grid cells with centroids inside this domain (plus
presence-rescued cells) form the final accessible grid:

```{r load-grid}
grid <- st_read(
  here("data", "processed", "grid", "ca_grid10km_accessible.gpkg"),
  quiet = TRUE
)

cat("Total accessible grid cells:", nrow(grid), "\n")
```

```{r map-grid-and-points, fig.width=5.5, fig.height=5.9}
pts_map <- st_transform(pts, st_crs(grid))
ca_map  <- st_transform(ca, st_crs(grid))

ggplot() +
  geom_sf(data = grid, fill = "gray80", color = "gray60", linewidth = 0.2) +
  geom_sf(data = ca_map, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts_map, color = "black", fill = "#2CBEC9",
          shape = 21, size = 1.7, stroke = 0.6, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "10 km accessible grid with bat observation locations",
    subtitle = "2,811 grid cells covering the accessible modeling domain"
  ) +
  theme_minimal(base_size = 13, base_family = "lato") +
  theme(
    plot.title = element_text(size = 13.5, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

The grid covers the portions of California where bat observations have
been recorded (plus a 50 km buffer), excluding remote areas like the far
eastern deserts and high Sierra peaks where no records exist. This
represents roughly 75% of the total California grid.

### 2.4 Landscape covariates

Three landscape variables were computed for each grid cell, each drawn
from a different remote sensing or geospatial dataset. The processing
pipeline (`R/04_build_grid_covariates.R`) follows a common pattern for
all raster covariates: crop the source raster to the grid extent,
reproject to California Albers (EPSG:3310), and extract a per-cell
summary statistic using `exactextractr::exact_extract()` or
`terra::extract()`.

-   **VIIRS nighttime light radiance** — annual composite images of
    nighttime light intensity (nW/cm²/sr) from the Earth Observation
    Group (VNL v2.1 for 2012–2021, v2.2 for 2022–2024). Per-year
    area-weighted means were averaged across all years to produce a
    single long-term mean radiance per cell.
-   **NLCD percent developed land** — the National Land Cover Database
    (2019, 30 m resolution) classifies pixels into land cover types. The
    proportion of pixels in the four "Developed" classes (Open Space
    through High Intensity) was computed per cell.
-   **GPWv4 population density** — the Gridded Population of the World
    v4 (rev. 11, \~1 km resolution) from NASA SEDAC, summarized per cell
    as an area-weighted mean.

A fourth covariate — percent protected land from PAD-US 4.1 (GAP Status
1–3) — was computed but excluded from the final modeling because it
added minimal predictive value once the urban variables were included.
The processing code is preserved in `R/04_build_grid_covariates.R`.

#### Combined covariate dataset

All three urban covariates were joined by `cell_id` into a single table.
Log-transformed versions of radiance and population density
(`log1p_radiance`, `log1p_pop_density`) were computed to reduce right
skew — both variables span several orders of magnitude, with most cells
near zero and a long tail of urban cells with very high values.

```{r load-covariates}
covariates <- read_csv(
  here("data", "processed", "covariates_grid", "grid_covariates_10km.csv"),
  show_col_types = FALSE
)

glimpse(covariates)
```

```{r covariate-summary}
covariates %>%
  summarise(
    across(c(mean_radiance, pct_developed, pop_density),
           list(Min = ~min(., na.rm = TRUE),
                Median = ~median(., na.rm = TRUE),
                Max = ~max(., na.rm = TRUE)),
           .names = "{.col}__{.fn}")
  ) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(variable = recode(variable,
    "mean_radiance" = "Nighttime radiance (nW/cm²/sr)",
    "pct_developed" = "Percent developed land",
    "pop_density"   = "Population density (people/km²)"
  ))
```

The ranges confirm what we'd expect: most cells have very low radiance,
little developed land, and sparse population, but a small number of
urban cells have extreme values across all three variables.

### 2.5 Building the final analysis dataset

The last step brings together the species presence data and the
landscape covariates into a single analysis-ready dataset.

**Why aggregate across years?** The landscape covariates vary far more
across space than over time at this scale, and 455 total observations
spread across 13 years and three species are far too sparse to support
year-level modeling. Aggregating across the full study period
(2012–2024) pools all available data into a single spatial snapshot,
maximizing statistical power at the cost of collapsing temporal dynamics
— a tradeoff revisited in the limitations section.

**Presence aggregation.** Each bat point was spatially joined to the
accessible grid (`st_join()` with `st_within()`) and aggregated by cell
and species. `tidyr::complete()` filled in every cell × species
combination with explicit zeros, and a binary `is_present` column was
derived (see `R/05_build_grid_presence.R` and
`R/06_build_grid_model_dataset.R` for details).

**Final merge.** The presence panel was joined to the covariate table by
`cell_id` using a left join, producing one row per cell × species with
both response and predictor variables:

```{r load-final-dataset}
df <- read_csv(
  here("data", "processed", "analysis_grid", "grid_model_dataset_10km.csv"),
  show_col_types = FALSE
)
```

```{r final-dataset-glimpse}
glimpse(df)
```

```{r final-dataset-summary}
df %>%
  summarise(
    n_rows = n(),
    n_cells = n_distinct(cell_id),
    n_species = n_distinct(species),
    total_observations = sum(n_obs),
    overall_presence_rate = round(mean(is_present), 4)
  )
```

The final dataset contains 8,433 rows (2,811 cells × 3 species), each
with three landscape covariates and their log-transformed versions. This
is the dataset used for all exploration and modeling that follows.

## 3. Exploratory data analysis

### 3.1 Response structure: extreme sparsity

```{r presence-absence-bar, fig.width=6.5, fig.height=3.9}
pres_summary <- df %>%
  group_by(species) %>%
  summarise(
    n_cells = n(),
    n_present = sum(is_present),
    prop_present = round(mean(is_present), 3),
    .groups = "drop"
  )

df %>%
  mutate(
    status = if_else(is_present == 1, "Present (\u22651 record)", "Absent (0 records)")
  ) %>%
  count(species, status) %>%
  group_by(species) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = species, y = prop, fill = status)) +
  geom_col() +
  geom_text(
    data = pres_summary,
    aes(x = species, y = prop_present, label = prop_present),
    inherit.aes = FALSE,
    hjust = -0.3, size = 5, fontface = "bold", color = "#1D9E75"
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Present (\u22651 record)" = "#1D9E75",
               "Absent (0 records)" = "gray75")
  ) +
  labs(
    title = "Most grid cells have no bat observations",
    subtitle = "Proportion of 2,811 accessible cells with at least one record (2012\u20132024)",
    x = NULL,
    y = "Proportion of cells",
    fill = NULL
  ) +
  theme_minimal(base_size = 13, base_family = "lato") +
  theme(legend.position = "top",
        plot.title = element_text(size = 15, face = "bold"),
        plot.subtitle = element_text(size = 11))
```

All three species are observed in fewer than 5% of grid cells.
*Aeorestes cinereus* is the most frequently observed, while *Myotis
californicus* is the rarest. This extreme sparsity motivates modeling
binary presence (observed or not) rather than observation counts — any
count-based model would be overwhelmed by structural zeros.

### 3.2 Mapping observations over the human-modified landscape

```{r prepare-map-data}
# Join covariates to the spatial grid for mapping
grid_map <- grid %>%
  mutate(cell_id = as.character(cell_id)) %>%
  left_join(
    df %>%
      mutate(cell_id = as.character(cell_id)) %>%
      distinct(cell_id, log1p_pop_density, log1p_radiance, pct_developed),
    by = "cell_id"
  )

# Ensure CRS alignment for all layers
pts_map <- st_transform(pts, st_crs(grid_map))
ca_map  <- st_transform(ca, st_crs(grid_map))
```

```{r map-pop-density, fig.width=6.5, fig.height=7}
# Define breaks in original units (roughly log-spaced)
pop_breaks <- c(0, 1, 10, 100, 1000, 5000, Inf)
pop_labels <- c("0 – 1", "1 – 10", "10 – 100", "100 – 1,000",
                "1,000 – 5,000", "> 5,000")

# Bin the log-transformed values using the transformed breaks
grid_map <- grid_map %>%
  mutate(
    pop_bin = cut(
      log1p_pop_density,
      breaks = log1p(pop_breaks),
      labels = pop_labels,
      include.lowest = TRUE,
      right = FALSE
    )
  )

# Viridis colors sampled at 6 evenly spaced points
bin_colors <- viridis::viridis(6, option = "cividis")

ggplot() +
  geom_sf(data = grid_map, aes(fill = pop_bin), color = NA) +
  scale_fill_manual(
    values = setNames(bin_colors, pop_labels),
    name = "Population density\n(people/km\u00B2)",
    na.value = "gray90",
    drop = FALSE
  ) +
  geom_sf(data = ca_map, fill = NA, color = "black", linewidth = 0.7) +
  geom_sf(data = pts_map, color = "black", fill = "#2CBEC9",
          shape = 21, size = 2.5, stroke = 0.6, alpha = 0.9) +
  coord_sf(datum = NA) +
  labs(
    title = "GBIF bat observations cluster near urban areas",
    subtitle = "Cyan points = bat observations; background = population density on accessible 10 km grid"
  ) +
  theme_minimal(base_size = 13, base_family = "lato") +
  theme(
    legend.position = "right",
    legend.key.height = unit(0.5, "cm"),
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 11.5)
  )
```

The map reveals a clear pattern: bat observations are concentrated in
and around California's major population centers — the San Francisco Bay
Area, Los Angeles basin, San Diego, and Sacramento. Large swaths of the
state's interior, mountains, and deserts have few or no records. This is
the observer bias problem in action: people (and therefore observation
effort) concentrate in cities.

Maps of nighttime light radiance and percent developed land (not shown)
display nearly identical spatial patterns — the same urban cores light
up, the same rural areas stay dark. This visual redundancy is confirmed
by the correlation analysis below.

### 3.3 Predictor correlations

```{r correlation-matrix}
cov_df <- df %>%
  select(log1p_radiance, log1p_pop_density, pct_developed) %>%
  drop_na()

cor_mat <- round(cor(cov_df), 3)
cor_mat
```

Nighttime light, population density, and percent developed land are all
strongly correlated (r = 0.76–0.87). At 10 km resolution, these
variables essentially capture the same underlying gradient of human
modification. Including all three in a single model would introduce
redundancy without adding predictive value — a finding that the model
comparison in the next section confirms formally.

### 3.4 Presence rates along the population density gradient

To explore how bat presence varies with urban intensity, we divide
population density into decile bins and compute the share of cells with
at least one observation in each bin, separately by species.

```{r species-colors}
# Consistent species color palette used throughout
species_colors <- c(
  "Aeorestes cinereus"  = "#1B9E77",
  "Myotis yumanensis"   = "#D95F02",
  "Myotis californicus" = "#7570B3"
)
```

```{r presence-vs-pop, fig.width=7.5, fig.height=5}
df %>%
  mutate(bin = ntile(log1p_pop_density, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_pop_density, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = species_colors) +
  theme_minimal(base_size = 13, base_family = "lato") +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 12),
  ) +
  labs(
    title = "Presence rate increases with population density",
    subtitle = "Binned deciles of log(1 + population density)",
    x = "Mean log(1 + population density) within bin",
    y = "Presence rate",
    color = "Species"
  )
```

Presence rates rise sharply in the upper deciles of population density,
and the three species diverge: *Aeorestes cinereus* shows the steepest
increase, while *Myotis californicus* responds more weakly. Equivalent
plots for nighttime radiance and percent developed land (not shown)
display the same qualitative pattern — unsurprising given the high
correlations among these predictors. These exploratory patterns motivate
the formal modeling that follows.

## 4. Modeling bat presence across the urban gradient

### 4.1 Why logistic regression?

The EDA revealed that bat presence is extremely sparse (fewer than 5% of
cells for any species) and that the response of interest is binary: was
a species ever observed in a given grid cell, or not? Logistic
regression is a natural fit for this kind of data. It models the
probability of presence as a function of predictor variables, using a
logit link to keep predicted probabilities between 0 and 1.

In notation, the model takes the form:

> log(p / (1 − p)) = β₀ + β₁ · predictor

where *p* is the probability of presence. The coefficient β₁ tells us
how the log-odds of presence change per unit increase in the predictor,
and exp(β₁) gives the **odds ratio** — the multiplicative change in odds
per unit increase.

All predictors were standardized (z-scored) before modeling so that
coefficients are comparable across variables measured on different
scales.

```{r standardize-predictors}
# Standardize predictors for comparable coefficients
df_model <- df %>%
  mutate(
    z_radiance  = scale(log1p_radiance)[, 1],
    z_pop       = scale(log1p_pop_density)[, 1],
    z_dev       = scale(pct_developed)[, 1],
    species     = factor(species)
  )
```

### 4.2 Comparing urban predictors

Given the strong collinearity among nighttime light, population density,
and percent developed land, a key question is: **does adding light or
development improve predictions beyond population density alone?**

To answer this, we fit a series of logistic regression models and
compared their predictive performance using 5-fold cross-validated
**AUC** (area under the ROC curve). AUC measures how well a model
distinguishes between presence and absence cells — a value of 0.5 means
no better than chance, and 1.0 means perfect discrimination. Models were
fit separately for each species.

```{r cv-function}
# Cross-validation helper: returns mean and SD of AUC across k folds
run_cv_model <- function(data, formula, k = 5) {
  folds <- rsample::vfold_cv(data, v = k, strata = is_present)

  results <- map_df(folds$splits, function(split) {
    train <- rsample::analysis(split)
    test  <- rsample::assessment(split)

    model <- glm(formula, data = train, family = binomial())
    probs <- predict(model, newdata = test, type = "response")
    auc_val <- pROC::auc(test$is_present, probs)

    tibble(auc = as.numeric(auc_val))
  })

  tibble(mean_auc = mean(results$auc), sd_auc = sd(results$auc))
}
```

```{r define-models}
model_list <- list(
  S1_pop       = is_present ~ z_pop,
  S2_radiance  = is_present ~ z_radiance,
  S3_dev       = is_present ~ z_dev,
  M1_pop_rad   = is_present ~ z_pop + z_radiance,
  M2_pop_dev   = is_present ~ z_pop + z_dev,
  M4_pop_rad_dev = is_present ~ z_pop + z_radiance + z_dev
)
```

```{r run-cv, cache=TRUE}
species_results <- df_model %>%
  group_split(species) %>%
  map_df(function(sp_data) {
    sp_name <- unique(sp_data$species)
    map_df(names(model_list), function(mname) {
      cv_res <- run_cv_model(sp_data, model_list[[mname]])
      tibble(
        species  = sp_name,
        model    = mname,
        mean_auc = cv_res$mean_auc,
        sd_auc   = cv_res$sd_auc
      )
    })
  })
```

```{r auc-barplot, fig.width=12, fig.height=5.45}
species_results %>%
  filter(model %in% c("S1_pop", "M1_pop_rad", "M2_pop_dev", "M4_pop_rad_dev")) %>%
  mutate(
    model = recode(model,
      "S1_pop"         = "Pop. density only",
      "M1_pop_rad"     = "+ nighttime light",
      "M2_pop_dev"     = "+ developed land",
      "M4_pop_rad_dev" = "+ light + development"
    ),
    model = factor(model, levels = c(
      "+ light + development",
      "+ developed land",
      "+ nighttime light",
      "Pop. density only"
    )),
    highlight = model == "Pop. density only"
  ) %>%
  ggplot(aes(x = mean_auc, y = model, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = sprintf("%.3f", mean_auc)),
    hjust = -0.15, size = 3.5
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#2C7BE5", "FALSE" = "gray50"),
    guide = "none"
  ) +
  facet_wrap(~ species) +
  coord_cartesian(xlim = c(0.5, 1.0)) +
  labs(
    title = "Population density alone is nearly as predictive as multi-variable models",
    subtitle = "5-fold cross-validated AUC by species",
    x = "Mean AUC",
    y = NULL
  ) +
  theme_minimal(base_size = 16, base_family = "lato") +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 17),
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "italic", size = 11),
    axis.line.x = element_line(color = "gray40", linewidth = 0.4),
    axis.line.y = element_line(color = "gray40", linewidth = 0.4),
    panel.spacing = unit(1.5, "lines"),
    panel.border = element_rect(color = "gray80", fill = NA, linewidth = 0.5),
    strip.background = element_rect(fill = "gray95", color = NA)
  )
```

Population density alone achieves cross-validated AUC scores of
approximately 0.76–0.80 depending on species. Adding nighttime light
radiance, percent developed land, or both improves AUC by less than 1
percentage point in every case. At 10 km resolution, these variables are
effectively measuring the same underlying gradient of human modification
— they are interchangeable as predictors of bat presence.

This does not mean light and development are ecologically irrelevant. It
means that at this spatial scale, with opportunistic data, their effects
cannot be separated from the broader urban intensity signal. For the
remaining analysis, we use **population density as the single urban
intensity predictor** — it is the most parsimonious, has the clearest
dual interpretation (accessibility proxy + human footprint), and
performs as well as any combination.

### 4.3 Do species respond differently?

Population density predicts bat presence, but do all three species
respond to it in the same way? To test this, we fit three nested
logistic models and compared them using **AIC** (Akaike Information
Criterion) — a measure of model fit that penalizes complexity. Lower AIC
indicates a better balance between fit and parsimony.

```{r interaction-model}
# Three nested models: pop only, additive species, species × pop interaction
model_pop         <- glm(is_present ~ z_pop,
                         data = df_model, family = binomial())
model_pop_species <- glm(is_present ~ z_pop + species,
                         data = df_model, family = binomial())
model_interaction <- glm(is_present ~ species * z_pop,
                         data = df_model, family = binomial())
```

```{r aic-comparison}
tibble(
  Model = c("Pop. density only",
            "Pop. + species (additive)",
            "Species \u00d7 pop. (interaction)"),
  Description = c(
    "Same slope and intercept for all species",
    "Different intercepts, same slope",
    "Different intercepts and slopes"
  ),
  AIC = c(AIC(model_pop),
          AIC(model_pop_species),
          AIC(model_interaction))
) %>%
  mutate(
    AIC = round(AIC, 1),
    delta_AIC = round(AIC - min(AIC), 1)
  )
```

The interaction model provides the best fit (lowest AIC), confirming
that species differ not just in their baseline presence rates but in the
strength of their association with population density. In other words,
the urban gradient does not affect all three species equally.

### 4.4 Species-specific effects

To interpret the interaction model, we extract the species-specific
slope (the effect of population density on log-odds of presence) and
convert it to an odds ratio with 95% confidence intervals.

```{r odds-ratios}
coefs    <- coef(model_interaction)
vcov_mat <- vcov(model_interaction)

# Helper: extract species-specific slope and standard error
get_species_slope <- function(sp) {
  if (sp == "Aeorestes cinereus") {
    slope <- coefs["z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"]
  } else {
    int_term <- paste0("species", sp, ":z_pop")
    slope <- coefs["z_pop"] + coefs[int_term]
    var   <- vcov_mat["z_pop", "z_pop"] +
             vcov_mat[int_term, int_term] +
             2 * vcov_mat["z_pop", int_term]
  }
  tibble(species = sp, log_odds_slope = slope, se = sqrt(var))
}

or_results <- bind_rows(
  get_species_slope("Aeorestes cinereus"),
  get_species_slope("Myotis californicus"),
  get_species_slope("Myotis yumanensis")
) %>%
  mutate(
    odds_ratio = round(exp(log_odds_slope), 2),
    lower_ci   = round(exp(log_odds_slope - 1.96 * se), 2),
    upper_ci   = round(exp(log_odds_slope + 1.96 * se), 2)
  )
```

```{r odds-ratio-table}
or_results %>%
  transmute(
    Species = species,
    `Odds ratio` = odds_ratio,
    `95% CI` = paste0("(", lower_ci, " – ", upper_ci, ")")
  )
```

An odds ratio greater than 1 means that presence probability increases
with population density. All three species show positive associations,
but the magnitudes differ: *Aeorestes cinereus* and *Myotis yumanensis*
have the strongest responses, while *Myotis californicus* shows a
notably weaker association with the urban gradient.

### 4.5 Predicted probability curves

The most interpretable way to visualize the interaction model is to plot
predicted presence probability across the full range of population
density, with the x-axis back-transformed to real-world units.

```{r predicted-curves, fig.width=7, fig.height=4.8}
# Back-transformation from z-score to population density
mu  <- mean(df_model$log1p_pop_density)
sig <- sd(df_model$log1p_pop_density)
z_to_pop <- function(z) exp((z * sig) + mu) - 1

# Build prediction grid
pred_grid <- expand.grid(
  z_pop   = seq(min(df_model$z_pop), max(df_model$z_pop), length.out = 200),
  species = levels(df_model$species)
)
pred_grid$pred_prob <- predict(
  model_interaction, newdata = pred_grid, type = "response"
)

# Reorder species for legend clarity
pred_grid$species <- factor(pred_grid$species, levels = c(
  "Aeorestes cinereus",
  "Myotis yumanensis",
  "Myotis californicus"
))

ggplot(pred_grid, aes(z_pop, pred_prob, color = species)) +
  geom_line(linewidth = 2) +
  scale_color_manual(values = species_colors) +
  scale_x_continuous(
    name = "Population density (people per km\u00B2, log-scaled)",
    labels = function(x) format(round(z_to_pop(x)), big.mark = ",")
  ) +
  labs(
    title = "Species differ in their association with the human-modified \nlandscape gradient",
    subtitle = "Predicted probability of presence from species \u00d7 population density interaction model",
    y = "Predicted probability of presence",
    color = "Species"
  ) +
  theme_minimal(base_size = 13, base_family = "lato") +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.line = element_line(linewidth = 0.8, color = "black"),
    panel.grid.major = element_line(linewidth = 0.5, color = "gray85"),
    panel.grid.minor = element_blank(),
    legend.position = c(0.25, 0.80),
    legend.background = element_rect(fill = "white", color = "gray85"),
    legend.text = element_text(face = "italic")
  )
```

This figure — the central result of the analysis — shows that:

-   **All three species** have higher predicted presence probability in
    more urbanized areas, consistent with observer bias driving the
    overall pattern.
-   ***Aeorestes cinereus*** shows the highest predicted probabilities
    across the gradient, reaching approximately 30% in the most densely
    populated cells.
-   ***Myotis californicus*** shows the weakest response — its predicted
    probability curve rises more slowly and remains lower than the other
    two species even at high population densities.
-   The **divergence between species** is the ecologically interesting
    signal. Even though observer bias inflates all three curves, the
    *relative differences* between species are informative: they suggest
    that *Myotis californicus* may be more sensitive to urbanization, or
    at least less associated with human-modified landscapes, than the
    other two species.

## 5. Discussion

### 5.1 Summary of findings

This analysis used 455 GBIF bat occurrence records across three focal
species, mapped onto a 10 km equal-area grid covering accessible regions
of California, to examine how components of the human-modified landscape
structure bat observation patterns. Three main findings emerged:

1.  **Population density alone is sufficient to predict bat presence at
    this scale.** Nighttime light radiance and percent developed land
    are strongly collinear with population density at 10 km resolution
    and add negligible predictive value (less than 1% AUC improvement).
    At this scale, these variables function as interchangeable urban
    intensity proxies.

2.  **All three species show positive associations with increasing urban
    intensity.** This is expected given the opportunistic nature of GBIF
    data — more people means more observers means more bat records,
    regardless of true ecological preference.

3.  **Species differ in the strength of their urban association.**
    *Myotis californicus* shows a significantly weaker association with
    population density than either *Aeorestes cinereus* or *Myotis
    yumanensis*. This species-level difference persists even after
    accounting for the shared effect of observer bias, suggesting a
    genuine ecological signal: *Myotis californicus* may be more
    sensitive to urbanization or less tolerant of human-modified
    landscapes.

### 5.2 The dual role of population density

Population density plays two roles in this analysis, and the tension
between them is worth making explicit. First, it serves as an
**accessibility proxy**: areas with more people have more observers, so
GBIF records are biased toward populated areas. Second, it serves as a
**human footprint measure**: population density genuinely correlates
with habitat modification, light pollution, noise, and other stressors
that affect bat ecology.

These roles cannot be fully separated with opportunistic data. The fact
that all three species show positive associations with population
density likely reflects both accessibility bias and some degree of
genuine urban tolerance. The more informative result is the *difference*
between species: if all three were equally biased by observer effort,
their curves would be parallel. The fact that *Myotis californicus*
diverges suggests something ecologically real.

### 5.3 Limitations and future directions

-   **Observer bias remains the dominant confound.** Without systematic
    survey data, it is not possible to determine whether species are
    truly more common near urban areas or simply more frequently
    reported there. Deploying acoustic monitoring arrays across the
    urban-to-wildland gradient would be the most direct way to
    disentangle ecological preference from reporting effort.

-   **Spatial autocorrelation is not modeled.** Nearby grid cells are
    likely more similar than distant ones, violating the independence
    assumption of standard logistic regression. Spatial random effects
    (e.g., via INLA) could account for this structure.

-   **Temporal dynamics are collapsed.** Aggregating across 2012–2024
    treats presence as static, which obscures potential range shifts or
    changes in observation effort. Multi-year monitoring data could
    reveal whether species are actively shifting in response to ongoing
    urbanization.

-   **Sample size is modest.** With 455 total records (and only 114 for
    *Myotis californicus*), effect estimates carry substantial
    uncertainty. Incorporating additional data sources (e.g.,
    iNaturalist, state wildlife agency databases) and expanding to more
    species would improve statistical power and test whether the urban
    filtering pattern holds across the broader bat community.

-   **Finer spatial resolution** (e.g., 1 km grids) combined with
    systematic survey data might allow nighttime light and land
    development to be separated as independent predictors, rather than
    collapsing into a single urban gradient as they do at 10 km.

### 5.4 So what? Why this matters

Despite these limitations, this analysis carries real implications for
how we think about bats, cities, and the data we use to study them.

**Opportunistic data is valuable — but it's not the whole picture.**
GBIF records can detect species-level differences in urban association,
proving that citizen science platforms are genuinely useful for
ecological research. But a species being "present" near a city doesn't
mean it's thriving there, and large portions of California have no bat
records at all — not because bats are absent, but because no one has
looked. Encouraging survey efforts in under-sampled regions would go a
long way toward separating real ecological patterns from artifacts of
observer effort.

**Cities may be filtering out sensitive species.** The divergence
between species hints at **urban filtering** — cities actively selecting
for species that tolerate or benefit from human infrastructure while
squeezing out those that can't. Some bats find advantages in urban
areas: buildings and bridges provide roost sites, and artificial lights
attract insect prey. Others — species that depend on dark forest or
natural roost structures — receive no such benefit and may be harmed by
light, noise, and fragmentation. The weaker urban association of *Myotis
californicus* is consistent with this process, and if urbanization is
quietly reshaping bat communities, that has real consequences for
biodiversity.

**Biodiversity data should inform urban planning.** As California
continues to grow, integrating species vulnerability data into land use
decisions could help identify where development is likely to have
outsized impacts on sensitive species — and where conservation buffers,
wildlife corridors, or dark-sky protections might make a meaningful
difference.

## 6. Data sources and citations

| Dataset | Source | Access |
|------------------------|------------------------|------------------------|
| Bat occurrence records | GBIF (2012–2024) | [gbif.org](https://www.gbif.org/) |
| Nighttime light radiance | VIIRS VNL v2.1/v2.2, EOG Payne Institute | [eogdata.mines.edu](https://eogdata.mines.edu/products/vnl/) |
| Land cover | NLCD 2019 | [mrlc.gov](https://www.mrlc.gov/) |
| Protected areas | PAD-US 4.1 | [usgs.gov/programs/gap-analysis-project](https://www.usgs.gov/programs/gap-analysis-project/science/pad-us-data-download) |
| Population density | GPWv4 rev. 11 (SEDAC/CIESIN) 2020 | [earthdata.nasa.gov](https://www.earthdata.nasa.gov/data/catalog/sedac-ciesin-sedac-gpwv4-popdens-r11-4.11) |
| State boundary | US Census TIGER/Line | via `tigris` R package |

```{r session-info}
sessionInfo()
```
