---
title: "01_EDA_sandbox_10km_grid"
output: html_notebook
---

```{r}
library(tidyverse)

# Read in the dataset
df <- readr::read_csv(
  here::here("data", "processed", "analysis_grid", "grid_model_dataset_10km.csv"),
  show_col_types = FALSE
)
```

```{r}
#Summary table of response
library(dplyr)

eda_presence_vs_count <- df %>%
  group_by(species) %>%
  summarise(
    n_rows = n(),                                # cell × species rows
    n_cells = n_distinct(cell_id),
    n_present = sum(is_present, na.rm = TRUE),    # occupied cell × species
    share_present = mean(is_present, na.rm = TRUE),
    n_zero_rows = sum(is_present == 0, na.rm = TRUE),
    share_zero_rows = mean(is_present == 0, na.rm = TRUE),
    total_obs = sum(n_obs, na.rm = TRUE),         # total GBIF records for species
    mean_n_obs_all = mean(n_obs, na.rm = TRUE),   # includes zeros
    mean_n_obs_given_present = mean(n_obs[is_present == 1], na.rm = TRUE),
    max_n_obs = max(n_obs, na.rm = TRUE),
    share_of_obs_in_top5_cells = {
      x <- sort(n_obs[is_present == 1], decreasing = TRUE)
      if (length(x) == 0) NA_real_ else sum(head(x, 5)) / sum(x)
    },
    .groups = "drop"
  ) %>%
  arrange(desc(share_present))

eda_presence_vs_count
```
```{r}
#Most rows are zeros
library(ggplot2)
library(tidyr)

df %>%
  mutate(present_label = if_else(is_present == 1, "Present (≥1 record)", "Absent (0 records)")) %>%
  count(species, present_label) %>%
  group_by(species) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = species, y = prop, fill = present_label)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most cell × species combinations are absences (zero-heavy data)",
    x = NULL,
    y = "Proportion of rows",
    fill = NULL
  ) +
  theme_minimal()
```



```{r}
#Distributions of predictor variables

vars <- c("log1p_radiance","log1p_pop_density","pct_developed","pct_protected")

df %>%
  select(all_of(vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~name, scales = "free") +
  theme_minimal()

df %>%
  summarise(
    share_radiance_zero = mean(log1p_radiance == 0, na.rm = TRUE),
    share_pop_zero = mean(log1p_pop_density == 0, na.rm = TRUE),
    share_dev_zero = mean(pct_developed == 0, na.rm = TRUE),
    share_prot_zero = mean(pct_protected == 0, na.rm = TRUE)
  )
```
```{r}
#Check correlation of predictor variables
#Collinearity: is “light” separable from “urbanization / accessibility”
cov_df <- df %>% 
  select(log1p_radiance, log1p_pop_density, pct_developed, pct_protected) %>%
  drop_na()

round(cor(cov_df), 3)
```
```{r}
#Bat observations on map
library(sf)
library(dplyr)
library(ggplot2)
library(scales)

# ---- Paths ----
grid_gpkg <- here::here("data", "processed", "grid", "ca_grid10km_accessible.gpkg")
pts_gpkg  <- here::here("data", "processed", "gbif", "gbif_bats_points_clean_CA_2012_2024.gpkg")
ca_gpkg   <- here::here("data", "processed", "boundaries", "ca_boundary.gpkg")

# ---- Read layers ----
grid <- st_read(grid_gpkg, quiet = TRUE)
pts  <- st_read(pts_gpkg, quiet = TRUE)
ca   <- st_read(ca_gpkg, layer = "ca_boundary", quiet = TRUE)

# ---- Join covariates from df ----
grid_map <- grid %>%
  mutate(cell_id = as.character(cell_id)) %>%
  left_join(
    df %>%
      mutate(cell_id = as.character(cell_id)) %>%
      distinct(cell_id, log1p_pop_density, log1p_radiance, pct_developed),
    by = "cell_id"
  )

# Align CRS
pts <- st_transform(pts, st_crs(grid_map))
ca  <- st_transform(ca, st_crs(grid_map))

# ---- Bright species colors ----
species_colors <- c(
  "Aeorestes cinereus"   = "#FF3B30",  # bright red
  "Myotis californicus"  = "#00E5FF",  # bright cyan
  "Myotis yumanensis"    = "#FFD60A"   # bright yellow
)

# ---- Map layers ----
bat_points <- geom_sf(
  data = pts,
  aes(color = species),
  size = 1.6,
  alpha = 0.9
)

ca_outline <- geom_sf(
  data = ca,
  fill = NA,
  color = "black",
  linewidth = 0.7
)

base_theme <- theme_minimal() +
  theme(
    panel.grid.major = element_line(linewidth = 0.2),
    legend.position = "right"
  )

# ====================================================
# Population Density Map
# ====================================================
p_pop <- ggplot() +
  geom_sf(
    data = grid_map,
    aes(fill = log1p_pop_density),
    color = NA
  ) +
  scale_fill_gradient(
    low = "gray90",
    high = "steelblue4",
    name = "log(1 + pop density)"
  ) +
  ca_outline +
  bat_points +
  scale_color_manual(values = species_colors) +
  coord_sf(datum = NA) +
  labs(
    title = "Bat observations over population density",
    subtitle = "Accessible 10 km grid"
  ) +
  base_theme

# ====================================================
# Light Radiance Map
# ====================================================
p_light <- ggplot() +
  geom_sf(
    data = grid_map,
    aes(fill = log1p_radiance),
    color = NA
  ) +
  scale_fill_gradient(
    low = "gray90",
    high = "steelblue4",
    name = "log(1 + radiance)"
  ) +
  ca_outline +
  bat_points +
  scale_color_manual(values = species_colors) +
  coord_sf(datum = NA) +
  labs(
    title = "Bat observations over nighttime light intensity",
    subtitle = "VIIRS radiance aggregated to 10 km grid"
  ) +
  base_theme

# ====================================================
# Developed Land Map
# ====================================================
p_dev <- ggplot() +
  geom_sf(
    data = grid_map,
    aes(fill = pct_developed),
    color = NA
  ) +
  scale_fill_gradient(
    low = "gray92",
    high = "steelblue4",
    name = "% developed",
    labels = percent_format(accuracy = 1)
  ) +
  ca_outline +
  bat_points +
  scale_color_manual(values = species_colors) +
  coord_sf(datum = NA) +
  labs(
    title = "Bat observations over developed land cover",
    subtitle = "NLCD developed classes aggregated to 10 km grid"
  ) +
  base_theme

# ---- Display maps ----
p_pop
p_light
p_dev
```


```{r}
#Association of Bat Presence vs. Pred Variables

#vs. Log Radiance
df %>%
  mutate(bin = ntile(log1p_radiance, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_radiance, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Presence rate vs light (binned deciles)",
    x = "Mean log(1 + radiance) within bin",
    y = "Presence rate"
  )



#vs. Log Pop Density
df %>%
  mutate(bin = ntile(log1p_pop_density, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(log1p_pop_density, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Presence rate vs population density (binned deciles)",
    x = "Mean log(1 + population density) within bin",
    y = "Presence rate"
  )

# vs. Percent Developed
df %>%
  mutate(bin = ntile(pct_developed, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(pct_developed, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Presence rate vs developed land (binned deciles)",
    x = "Mean percent developed within bin",
    y = "Presence rate"
  )

# vs. Percent Protected
df %>%
  mutate(bin = ntile(pct_protected, 10)) %>%
  group_by(species, bin) %>%
  summarise(
    x = mean(pct_protected, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p, color = species)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Presence rate vs protected land (binned deciles)",
    x = "Mean percent protected within bin",
    y = "Presence rate"
  )
```

```{r}
df %>%
  mutate(
    pop_bin = ntile(log1p_pop_density, 10),
    light_bin = ntile(log1p_radiance, 10)
  ) %>%
  group_by(species, pop_bin, light_bin) %>%
  summarise(
    p = mean(is_present, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(pop_bin, light_bin, fill = p)) +
  geom_tile() +
  facet_wrap(~species) +
  theme_minimal() +
  labs(
    title = "Presence rate across accessibility (pop) and light gradients",
    x = "Pop density decile",
    y = "Radiance decile",
    fill = "Presence rate"
  )
```
```{r}
df %>%
  mutate(pop_q = ntile(log1p_pop_density, 4),
         prot_bin = ntile(pct_protected, 10)) %>%
  group_by(species, pop_q, prot_bin) %>%
  summarise(
    x = mean(pct_protected, na.rm = TRUE),
    p = mean(is_present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x, p)) +
  geom_line() +
  geom_point() +
  facet_grid(species ~ pop_q) +
  theme_minimal() +
  labs(
    title = "Presence vs protected land, stratified by pop density quartile",
    x = "Mean pct protected",
    y = "Presence rate"
  )
```
Based on EDA, pick:

Core predictors: usually log1p_pop_density, pct_developed, log1p_radiance

Secondary: pct_protected

Decide: interactions by species? nonlinear?

A clean candidate set:

M0: is_present ~ species (baseline)

M1: is_present ~ species + log1p_pop_density

M2: ... + pct_developed

M3: ... + log1p_radiance

M4: ... + pct_protected

M5: species * (log1p_radiance + pct_developed + log1p_pop_density + pct_protected) (full interaction)

If EDA suggests curvature: swap glm for mgcv::gam later.



