---
title: "01_master_analysis"
output: html_notebook
---

## Prerequisites

This analysis notebook assumes that all data acquisition and processing scripts
in the `R/` directory have already been executed.

Specifically, the following scripts should be run **once, in order**, prior to
rendering this notebook:

1. `R/01_get_gbif_bats.R` — download raw GBIF bat occurrence data (2012–present)
2. `R/02_get_ca_counties.R` — acquire California county boundaries
3. `R/03_process_viirs_county_year.R` — summarize annual VIIRS nighttime lights to county × year
4. `R/04_clean_gbif_to_ca.R` — spatially filter GBIF occurrences to California
5. `R/05_aggregate_bats_county_year.R` — aggregate bat counts to county × year × species

These scripts write processed datasets to the `data/processed/` directory, which
are read by this notebook.

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(scales)
library(here)
library(viridis)
library(broom)
library(purrr)
```
```{r}
# Load analysis-ready dataset (county × year × species)
dat <- readr::read_csv(
  here("data/processed/analysis/bats_viirs_county_year_species.csv"),
  show_col_types = FALSE
)


# Basic derived variables used throughout
dat <- dat %>%
  mutate(
    is_zero = as.integer(n_obs == 0),
    is_present = as.integer(n_obs > 0)  # presence indicator for later modeling
  )
```

This analysis uses an aggregated county–year–species dataset combining GBIF bat occurrence counts (n_obs) and VIIRS nighttime light intensity (mean_radiance, log1p_radiance). Because GBIF records are opportunistic, n_obs is interpreted as reporting intensity rather than abundance. The primary goal is to assess whether nighttime light is associated with the probability of reporting any bats, and whether that relationship differs by species.

## 1) Response structure: sparsity and zeros
### 1.1 Summary statistics and zero share


```{r}
dat %>%
  summarise(
    n_rows = n(),
    mean_n_obs = mean(n_obs),
    var_n_obs = var(n_obs),
    share_zero = mean(n_obs == 0),
    share_nonzero = mean(n_obs > 0),
    max_n_obs = max(n_obs)
  )

dat %>%
  group_by(species_label) %>%
  summarise(
    mean_n_obs = mean(n_obs),
    share_zero = mean(n_obs == 0),
    max_n_obs = max(n_obs),
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))

```

Observation counts are extremely sparse: most county–year–species combinations contain zero reported observations. This motivates treating the process as primarily presence (zero vs non-zero) rather than intensity, and suggests that simple Poisson count models are inappropriate due to strong overdispersion and zero inflation.

### 1.2 Distribution of Observation Counts by Species

```{r}
ggplot(dat, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Distribution of bat observation counts by species",
    subtitle = "Most county–year–species rows contain zero observations",
    x = "Observation count (n_obs)",
    y = "Number of county–year–species rows"
  ) +
  theme_minimal()

```

Each species shows strong zero inflation. Even among nonzero rows, counts are typically small, reinforcing that the most stable signal is whether observations occur at all.

## 2) Spatial context and heterogeneity
### 2.1 County-level share of zero observations 

```{r}
# Read county boundaries (produced earlier in your pipeline)
ca_counties_sf <- st_read(
  here("data", "processed", "boundaries", "ca_counties.gpkg"),
  quiet = TRUE
)

county_zero <- dat %>%
  group_by(GEOID, county_name) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    share_present = mean(n_obs > 0),
    total_obs = sum(n_obs),
    .groups = "drop"
  )

county_zero_map <- ca_counties_sf %>%
  left_join(county_zero, by = "GEOID")

ggplot(county_zero_map) +
  geom_sf(aes(fill = share_zero), color = "white", linewidth = 0.2) +
  scale_fill_viridis(option = "magma", labels = percent, name = "Share zero") +
  labs(
    title = "Spatial heterogeneity in reporting: share of zero-observation rows by county",
    subtitle = "Higher values indicate persistent sparsity across species and years"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```
Zeros are not randomly distributed in space: many counties exhibit persistently high zero shares (“data deserts”), while a smaller set of counties show consistent reporting. This highlights strong spatial heterogeneity in sampling effort and/or habitat suitability, motivating cautious interpretation of bat–light associations. 

## 3) Light (VIIRS) structure: mostly spatial, weakly temporal
### 3.1 Between-county vs within-county radiance variance (numeric justification)

County-level nighttime radiance is highly right-skewed, with most California counties exhibiting very low mean radiance and a small number of highly urbanized counties forming a long, bright tail (see Appendix Figure 1). All analyses therefore use a log(1 + radiance) transformation

```{r}
county_mean_radiance <- dat %>%
  group_by(GEOID) %>%
  summarise(mean_log_radiance = mean(log1p_radiance, na.rm = TRUE), .groups = "drop")

var_between <- var(county_mean_radiance$mean_log_radiance, na.rm = TRUE)

var_within <- dat %>%
  group_by(GEOID) %>%
  summarise(v = var(log1p_radiance, na.rm = TRUE), .groups = "drop") %>%
  summarise(mean_within_var = mean(v, na.rm = TRUE)) %>%
  pull(mean_within_var)

tibble(var_between = var_between, mean_within_var = var_within)

```
```{r}
dat %>%
  group_by(year) %>%
  summarise(mean_radiance_state = mean(mean_radiance, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(year, mean_radiance_state)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Statewide mean nighttime radiance over time",
    subtitle = "Radiance varies modestly across years compared to differences between counties",
    x = "Year",
    y = "Mean radiance"
  ) +
  theme_minimal()

```
Nighttime light intensity is largely stable over time at the county scale: between-county differences dominate within-county year-to-year variation. This supports interpreting radiance primarily as persistent spatial context, and motivates a modeling approach focused on spatial associations rather than temporal causality.

## 4) Core EDA result: the zero process depends on radiance and species
Aggregated over years, total observation counts are weakly correlated with county-level radiance, but this relationship is dominated by a small number of high-reporting counties and collapses under sensitivity checks. For this reason, subsequent analysis focuses on presence rather than cumulative counts.

### 4.1 Exploratory “zero probability vs radiance”

```{r}
ggplot(dat, aes(log1p_radiance, is_zero)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial),
    se = TRUE
  ) +
  facet_wrap(~ species_label) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Exploratory relationship: probability of zero observations vs radiance",
    subtitle = "Each facet uses a binomial smoother; interpret as descriptive, not inferential",
    x = "log(1 + mean radiance)",
    y = "Pr(n_obs = 0)"
  ) +
  theme_minimal()

```
To visualize the presence/absence structure directly, I model is_zero = 1(n_obs = 0). For a 0/1 indicator, the conditional mean equals the conditional probability:
\[
E[1(n_{obs}=0)\mid x] = P(n_{obs}=0 \mid x)
\]
The smoothed curves therefore provide a descriptive estimate of how the probability of observing no bats varies with radiance.

Across all county–year rows, higher radiance is associated with lower zero probability, but the strength of this relationship differs sharply by species. Aeorestes cinereus shows the strongest decline in zero probability with increasing radiance, while Myotis species remain highly zero-inflated across the radiance gradient.

### 4.2 Sensitivity check: does the zero–radiance relationship reflect coverage vs persistence?

```{r}
#Condition on counties that ever report each species

ever_obs_counties <- dat %>%
  group_by(species_label, GEOID, county_name) %>%
  summarise(ever_obs = any(n_obs > 0), .groups = "drop") %>%
  filter(ever_obs)

dat_ever <- dat %>%
  inner_join(ever_obs_counties, by = c("species_label", "GEOID", "county_name"))


#Plot zero probability vs radiance (ever-observed only)
ggplot(dat_ever, aes(log1p_radiance, is_zero)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial),
    se = TRUE
  ) +
  facet_wrap(~ species_label) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Sensitivity check: zero probability vs radiance (ever-observed counties only)",
    subtitle = "Removes counties that never report a species; isolates persistence within the reporting network",
    x = "log(1 + mean radiance)",
    y = "Pr(n_obs = 0)"
  ) +
  theme_minimal()


```
Conditioning on counties where each species has ever been observed separates two mechanisms:
(1) coverage (which counties ever report bats at all) vs
(2) persistence (given coverage, whether bats are observed in a given year).
After conditioning, the radiance–zero relationship weakens substantially for Myotis species, suggesting radiance primarily affects reporting *coverage*. In contrast, Aeorestes cinereus retains a strong decline in zero probability with radiance, indicating that light is associated with repeated observations even within counties where the species is known to occur.

## 5) Modeling Plan
The EDA and sensitivity analyses indicate that the dominant, most stable signal in this dataset is the probability of observing any bats at all, not the intensity of observations once present. Zeros are structured, spatially heterogeneous, and strongly species-dependent. Nighttime radiance is largely a stable spatial covariate and shows the clearest association with presence/absence, particularly for Aeorestes cinereus.

Therefore, the next step is a species-stratified logistic regression modeling bat presence (n_obs > 0) as a function of log-transformed radiance (log1p_radiance), providing interpretable effect sizes and predicted presence probabilities by species.


## 6) Species-stratified logistic regression.


```{r}
# Split data by species
species_list <- split(dat, dat$species_label)

# Fit one logistic regression per species
fits <- map(
  species_list,
  ~ glm(is_present ~ log1p_radiance,
        family = binomial,
        data = .x)
)

fits


```

```{r}
coef_tbl <- map_dfr(
  names(fits),
  function(sp) {
    broom::tidy(fits[[sp]], conf.int = TRUE, exponentiate = TRUE) %>%
      filter(term == "log1p_radiance") %>%
      mutate(species_label = sp)
  }
)

coef_tbl %>%
  select(species_label, estimate, conf.low, conf.high, p.value)

```


```{r}
#Create a prediction grid
pred_df <- map_dfr(
  names(fits),
  function(sp) {

    # Radiance range for this species
    sp_dat <- dat %>% filter(species_label == sp)

    x_seq <- seq(
      quantile(sp_dat$log1p_radiance, 0.02),
      quantile(sp_dat$log1p_radiance, 0.98),
      length.out = 200
    )

    # Predict on link scale
    preds <- predict(
      fits[[sp]],
      newdata = tibble(log1p_radiance = x_seq),
      type = "link",
      se.fit = TRUE
    )

    tibble(
      species_label = sp,
      log1p_radiance = x_seq,
      eta = preds$fit,
      se = preds$se.fit
    )
  }
) %>%
  mutate(
    prob = plogis(eta),
    prob_low = plogis(eta - 1.96 * se),
    prob_high = plogis(eta + 1.96 * se)
  )

```

```{r}
#Plot predictions

ggplot(pred_df, aes(log1p_radiance, prob)) +
  geom_ribbon(aes(ymin = prob_low, ymax = prob_high), alpha = 0.2) +
  geom_line(linewidth = 1) +
  facet_wrap(~ species_label) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Predicted probability of bat presence vs nighttime radiance",
    subtitle = "Species-stratified logistic regression (95% confidence bands)",
    x = "log(1 + mean radiance)",
    y = "Predicted Pr(n_obs > 0)"
  ) +
  theme_minimal()

```

```{r}
ggplot(coef_tbl, aes(x = estimate, y = species_label)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_x_log10() +
  labs(
    title = "Radiance effect on bat presence by species",
    subtitle = "Odds ratios for 1-unit increase in log(1 + radiance)",
    x = "Odds ratio (log scale)",
    y = NULL
  ) +
  theme_minimal()

```





## Appendix: Additional exploratory analyses

This appendix contains exploratory figures used to understand data structure but not included in the main analysis to maintain narrative focus.

```{r}
#Distribution of counties by radiance level

county_radiance <- dat %>%
  group_by(GEOID, county_name) %>%
  summarise(
    mean_radiance = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(county_radiance, aes(mean_radiance)) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of mean nighttime radiance across counties",
    x = "Mean radiance (nW/cm²/sr)",
    y = "Number of counties"
  )

ggplot(county_radiance, aes(log1p(mean_radiance))) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of county radiance (log scale)",
    x = "log(1 + mean radiance)",
    y = "Number of counties"
  )
```

