---
title: "02_model_testing"
output: html_notebook
---

#Exploring how different aspects of human footprint 

```{r}
library(dplyr)
library(purrr)
library(rsample)
library(pROC)
library(broom)
library(tidyr)
library(stringr)

# Read in the dataset
df <- readr::read_csv(
  here::here("data", "processed", "analysis_grid", "grid_model_dataset_10km.csv"),
  show_col_types = FALSE
)

# Standardize predictors (z-score)
df_model <- df %>%
  mutate(
    z_radiance   = scale(log1p_radiance)[,1],
    z_pop        = scale(log1p_pop_density)[,1],
    z_dev        = scale(pct_developed)[,1],
    z_protected  = scale(pct_protected)[,1]
  )
```

```{r}
run_cv_model <- function(data, formula, k = 5) {
  
  folds <- vfold_cv(data, v = k, strata = is_present)
  
  results <- map_df(folds$splits, function(split) {
    
    train <- analysis(split)
    test  <- assessment(split)
    
    model <- glm(formula,
                 data = train,
                 family = binomial())
    
    # Predict probabilities
    probs <- predict(model, newdata = test, type = "response")
    
    # Compute AUC
    auc_val <- pROC::auc(test$is_present, probs)
    
    tibble(auc = as.numeric(auc_val))
  })
  
  tibble(
    mean_auc = mean(results$auc),
    sd_auc   = sd(results$auc)
  )
}
```



```{r}
model_list <- list(
  
  # ---- Single-predictor models ----
  S1_pop        = is_present ~ z_pop,
  S2_radiance   = is_present ~ z_radiance,
  S3_dev        = is_present ~ z_dev,
  S4_protected  = is_present ~ z_protected,
  
  # ---- Population + one additional predictor ----
  M1_pop_rad    = is_present ~ z_pop + z_radiance,
  M2_pop_dev    = is_present ~ z_pop + z_dev,
  M3_pop_prot   = is_present ~ z_pop + z_protected,
  
  # ---- Full urban footprint combinations ----
  M4_pop_rad_dev      = is_present ~ z_pop + z_radiance + z_dev,
  M5_full_no_interact = is_present ~ z_pop + z_radiance + z_dev + z_protected
)
```



```{r}
suppressMessages(species_results <- df_model %>%
  group_split(species) %>%
  map_df(function(sp_data) {
    
    sp_name <- unique(sp_data$species)
    
    map_df(names(model_list), function(mname) {
      
      cv_res <- run_cv_model(sp_data, model_list[[mname]])
      
      tibble(
        species = sp_name,
        model   = mname,
        mean_auc = cv_res$mean_auc,
        sd_auc   = cv_res$sd_auc
      )
    })
  }))
```

```{r}

#Species result by highest to lowest AUC
species_results %>%
  arrange(species, desc(mean_auc))

#Species result- does adding additional factors improve performance after accounting for pop density?
species_results %>%
  filter(model %in% c(
    "S1_pop",
    "M1_pop_rad",
    "M2_pop_dev",
    "M4_pop_rad_dev"
  )) %>%
  mutate(
    model = factor(
      model,
      levels = c(
        "S1_pop",
        "M1_pop_rad",
        "M2_pop_dev",
        "M4_pop_rad_dev"
      )
    )
  ) %>%
  arrange(species, model)


#AUC is the probability that a randomly chosen presence cell receives higher predicted probability than a randomly chosen absence cell
#Model predicts Myotis yumanensis slightly better than the other species
```

Variables Conclusion: At 10 km resolution, light, development, and protection do not add substantial predictive value once population density is included.

Now: Do species differ in their response to the human-modified landscape gradient?

```{r}
#Does including species improve prediction beyond population density alone?
#is_present ~ z_pop

model_pop <- glm(
  is_present ~ z_pop,
  data = df_model,
  family = binomial()
)
summary(m_pop)

#Add species:
#is_present ~ z_pop + species
#pop density has same effect for all species

model_pop_species <- glm(
  is_present ~ z_pop + species,
  data = df_model,
  family = binomial()
)
summary(m_species)


#Does the relationship between presence probability and population density differ among species?
# Make species a factor
#is_present∼species + z_pop + species∗z_pop
#species can have different slopes

df_model <- df_model %>%
  mutate(species = factor(species))

# Fit interaction model
model_interaction <- glm(
  is_present ~ species * z_pop,
  data = df_model,
  family = binomial()
)

summary(model_interaction)


#Model fit comparison
# Create comparison table
model_comparison <- tibble(
  Model = c("Pop only",
            "Pop + species",
            "Species × pop"),
  AIC = c(
    AIC(model_pop),
    AIC(model_pop_species),
    AIC(model_interaction)
  )
)

model_comparison
```

```{r}
# Tidy model output
coefs <- tidy(model_interaction)

# Extract coefficients manually for clarity
b0  <- coefs$estimate[coefs$term == "(Intercept)"]
b_sp_cal <- coefs$estimate[coefs$term == "speciesMyotis californicus"]
b_sp_yum <- coefs$estimate[coefs$term == "speciesMyotis yumanensis"]

b_pop <- coefs$estimate[coefs$term == "z_pop"]
b_int_cal <- coefs$estimate[coefs$term == "speciesMyotis californicus:z_pop"]
b_int_yum <- coefs$estimate[coefs$term == "speciesMyotis yumanensis:z_pop"]

# Build clean summary table
summary_table <- tibble(
  species = c("Aeorestes cinereus",
              "Myotis californicus",
              "Myotis yumanensis"),
  
  intercept_logodds = c(
    b0,
    b0 + b_sp_cal,
    b0 + b_sp_yum
  ),
  
  slope_logodds = c(
    b_pop,
    b_pop + b_int_cal,
    b_pop + b_int_yum
  )
) %>%
  mutate(
    odds_ratio_per_1SD = exp(slope_logodds),
    baseline_probability = plogis(intercept_logodds)
  )

summary_table
```



```{r}
#Plot predicted presence probability vs population density (standardized)
library(ggplot2)

# Create prediction grid
pred_grid <- expand.grid(
  z_pop = seq(min(df_model$z_pop),
              max(df_model$z_pop),
              length.out = 100),
  species = levels(df_model$species)
)

pred_grid$pred_prob <- predict(
  model_interaction,
  newdata = pred_grid,
  type = "response"
)

ggplot(pred_grid, aes(z_pop, pred_prob, color = species)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(
    title = "Predicted presence probability vs human-modification gradient",
    x = "Standardized population density (urban gradient)",
    y = "Predicted probability of presence"
  )



#Plot with intuitive x values
# Save the mean and sd used in standardization
mu  <- mean(df_model$log1p_pop_density)
sig <- sd(df_model$log1p_pop_density)

# Create prediction grid
pred_grid <- expand.grid(
  z_pop = seq(min(df_model$z_pop),
              max(df_model$z_pop),
              length.out = 100),
  species = levels(df_model$species)
)

pred_grid$pred_prob <- predict(
  model_interaction,
  newdata = pred_grid,
  type = "response"
)

# Function to convert z_pop back to population density
z_to_pop <- function(z){
  exp((z * sig) + mu) - 1
}

ggplot(pred_grid, aes(z_pop, pred_prob, color = species)) +
  geom_line(size = 1.2) +
  scale_x_continuous(
    name = "Population density (people per km², log scaled)",
    labels = function(x) round(z_to_pop(x), 0)
  ) +
  theme_minimal() +
  labs(
    title = "Predicted presence probability vs human-modified landscape gradient",
    y = "Predicted probability of presence"
  )
```

```{r}
# Extract coefficients and variance-covariance matrix
coefs <- coef(interaction_model)
vcov_mat <- vcov(interaction_model)

# Helper function to compute slope + SE for each species
get_species_slope <- function(species_name) {
  
  if (species_name == "Aeorestes cinereus") {
    slope <- coefs["z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"]
  }
  
  if (species_name == "Myotis californicus") {
    slope <- coefs["z_pop"] + coefs["speciesMyotis californicus:z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"] +
             vcov_mat["speciesMyotis californicus:z_pop",
                      "speciesMyotis californicus:z_pop"] +
             2 * vcov_mat["z_pop",
                          "speciesMyotis californicus:z_pop"]
  }
  
  if (species_name == "Myotis yumanensis") {
    slope <- coefs["z_pop"] + coefs["speciesMyotis yumanensis:z_pop"]
    var   <- vcov_mat["z_pop", "z_pop"] +
             vcov_mat["speciesMyotis yumanensis:z_pop",
                      "speciesMyotis yumanensis:z_pop"] +
             2 * vcov_mat["z_pop",
                          "speciesMyotis yumanensis:z_pop"]
  }
  
  se <- sqrt(var)
  
  tibble(
    species = species_name,
    log_odds_slope = slope,
    se = se
  )
}

# Build table
species_list <- c("Aeorestes cinereus",
                  "Myotis californicus",
                  "Myotis yumanensis")

results <- bind_rows(lapply(species_list, get_species_slope)) %>%
  mutate(
    odds_ratio = exp(log_odds_slope),
    lower_ci = exp(log_odds_slope - 1.96 * se),
    upper_ci = exp(log_odds_slope + 1.96 * se)
  )

results

ggplot(results, aes(x = species, y = odds_ratio)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci),
                width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Effect of Urban Intensity on Bat Presence",
    subtitle = "Odds ratio per 1 SD increase in population density",
    x = "Species",
    y = "Odds Ratio (95% CI)"
  )
```

Species Differences Key takeaway: Bat species differed in their response to the human-modified landscape gradient. While all three species showed positive associations with increasing urban intensity, Aeorestes cinereus and Myotis yumanensis exhibited substantially stronger increases in odds of presence per unit increase in population density than Myotis californicus, indicating greater urban association. Although Myotis yumanensis had a slightly larger multiplicative odds response, predicted probability curves appear highest for Aeorestes because it has a higher baseline probability of presence at average urban intensity. As a result, similar slopes in log-odds space translate into higher absolute probabilities for Aeorestes at the urban extreme. Together, these patterns suggest species-specific tolerance to human-modified landscapes and potential restructuring of bat community composition along urban gradients.
***important caveat: Because GBIF data are opportunistic, population density may capture both ecological association and reporting effort; therefore, these differences should be interpreted as differences in observed association along a human-activity gradient rather than definitive causal effects.***

Complete narrative outline:
My question: What aspects of human-modified landscapes structure bat presence at fine spatial scales, and do species respond differently to increasing urban intensity?

This study uses opportunistic biodiversity data aggregated to a 10 km equal-area grid and incorporates population density as an accessibility proxy to examine how human-modified landscapes structure bat presence patterns and whether species differ in their association with urban intensity.

--Population density (urban intensity proxy) is a strong predictor of bat presence at 10 km resolution.

--Artificial light and percent developed land do not add meaningful predictive value once population density is included. All 3 are very correlated and function well as urban intensity proxies.Percent protected less so.

--At this spatial scale, light and development largely reflect a broader human-modified landscape gradient rather than independent effects.

--All three species show positive associations with increasing urban intensity (used pop density).

--Aeorestes cinereus and Myotis yumanensis respond more strongly to the urban gradient than Myotis californicus.

--These results suggest species-specific differences in association with human-modified landscapes, though patterns reflect observed presence in opportunistic data rather than confirmed causal effects.