---
title: "eda-sandbox"
output: html_notebook
---

### Distribution of the Response (n_obs)

```{r}
#Load Packages and Data

library(tidyverse)
library(janitor)
library(ggplot2)
library(sf)
library(scales)   
library(ggrepel)
library(viridis)


data <- readr::read_csv(
  here::here(
    "data/processed/analysis/bats_viirs_county_year_species.csv"
  ),
  show_col_types = TRUE
)
```
```{r}
#Check out data

glimpse(data)

summary(data$n_obs)

```

All the zeroes bring the means number of observations per species-year-county down to only 0.2. Max number of observations in any one row is 19.

```{r}
#Proportion of zeros overall

data %>%
  summarise(
    share_zero = mean(n_obs == 0),
    share_nonzero = mean(n_obs > 0)
  )

#Counts of each species

data %>%
  group_by(species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  )

#Zeros by species

data %>%
  group_by(species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    mean_n_obs = mean(n_obs),
    max_n_obs  = max(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))
```

About 90% of all rows have 0 observations, indicating sparsity and zero inflation.
Aeorestes (hoary bat) has about 85% share of 0 observations while Myotis has 93-94%. This could say something about the difference in ease of detecting those species.

The majority of species–county–year combinations contain zero reported observations. This sparsity reflects the opportunistic nature of GBIF records, coarse spatial aggregation, and imperfect detection of nocturnal bats rather than true absence. As a result, observation counts are interpreted as relative reporting intensity rather than abundance.

```{r}
#Histogram of n_obs (raw scale)
ggplot(data, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  labs(
    title = "Distribution of bat observation counts",
    x = "Number of observations (n_obs)",
    y = "Number of species–county–year rows"
  )

#Histogram on log scale (to see tail structure)
ggplot(data %>% filter(n_obs > 0), aes(n_obs)) +
  geom_histogram(binwidth = 1, color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of nonzero bat observations (log scale)",
    x = "n_obs (log scale)",
    y = "Count"
  )


#Variance vs mean
data %>%
  summarise(
    mean_n_obs = mean(n_obs),
    var_n_obs  = var(n_obs)
  )


```
The distribution of nonzero observation counts is heavily right-skewed (small but influential tail!). Most species–county–year combinations with reported observations have only 1–3 records, while a small number exhibit substantially higher counts (up to ~30). This long-tailed structure indicates strong overdispersion and suggests that a small subset of counties and years account for a disproportionate share of reported bat activity.

Also, mean observations are 0.20 and variance of observations is 0.89. Variance > mean so that is overdispersion.Te small tail dominates the variance


```{r}
#Response distribution by species (faceted)
ggplot(data, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Distribution of observation counts by species",
    x = "n_obs",
    y = "Count"
  )
```

Observation counts are highly overdispersed relative to a Poisson distribution (mean ≈ 0.2, variance ≈ 0.89). While most nonzero counts are small (1–3 observations), a small number of county–year–species combinations exhibit moderately large counts, producing a long right tail. This pattern is most pronounced for Aeorestes cinereus, which shows both fewer zeros and higher maximum counts than the Myotis species.

The Poisson distribution is inappropriate for this dataset because observation counts are highly overdispersed relative to the mean, contain excess zeros, and reflect unmodeled heterogeneity in observation effort and habitat suitability across counties, years, and species.


### Temporal Structure

```{r}
#Total observations per year (overall)

data %>%
  group_by(year) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total reported bat observations per year",
    x = "Year",
    y = "Total observations"
  )

#Observations per year by species

data %>%
  group_by(year, species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total bat observations per year by species",
    x = "Year",
    y = "Total observations",
    color = "Species"
  )

```
I wonder what caused the peak in 2016?
In the earlier years, Myotis observations outnumber Aeroestes but by 2020, Aeroestes observations are much higher.




```{r}
#Number of counties with ≥1 observation per year
data %>%
  group_by(year) %>%
  summarise(
    n_counties_with_obs = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties_with_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of counties with ≥1 bat observation per year",
    x = "Year",
    y = "Counties reporting bats"
  )

#Species-specific county coverage over time
data %>%
  group_by(year, species_label) %>%
  summarise(
    n_counties = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "County coverage by species over time",
    x = "Year",
    y = "Counties with ≥1 observation",
    color = "Species"
  )
```
There  is a lot of fluctuation from year to year but overall appears to be a general trend towards spatial expansion of bat reporting.
Aeroestes had an increase in how widely they're reported over time, while the Myotis species both decreased, meaning the overall spatial expansion of bat reporting was driven by Aeroestes


```{r}
#Zero share over time- are zeros becoming less common?

data %>%
  group_by(year) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows over time",
    x = "Year",
    y = "Percent zero"
  )

#Zero share by species over time

data %>%
  group_by(year, species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero, color = species_label)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Zero-observation share over time by species",
    x = "Year",
    y = "Percent zero",
    color = "Species"
  )


```
Share of zeros over time has had a few percentage jumps but seems to have stabled around 90% in recent years.

Summary: Bat observation data exhibit strong temporal heterogeneity rather than a simple trend. Total observations peak sharply in 2016, driven primarily by Myotis californicus, followed by a decline and lower-level fluctuations in subsequent years. In contrast, Aeorestes cinereus observations become increasingly dominant after 2017, accounting for a growing share of both total observations and spatial coverage. While total counts decline after 2016, the number of counties with at least one reported observation increases over time, indicating broader but shallower reporting in recent years. High proportions of zero-observation rows persist throughout the study period, particularly for Myotis species, highlighting structural sparsity and species-specific detection dynamics.



### Spatial Structure


```{r}
#Total observations by county (summed over years)

county_totals <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(total_obs))

county_totals


ggplot(county_totals, aes(reorder(county_name, total_obs), total_obs)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Total bat observations by county (2012–2024)",
    x = "County",
    y = "Total observations"
  )


#Counties dominating total observations ("super-observer" counties)
#Cumulative contribution
county_totals %>%
  mutate(
    cumulative_obs = cumsum(total_obs),
    cumulative_share = cumulative_obs / sum(total_obs)
  )




```

```{r}
# ---- Map: total observations by county ----


# Read CA counties geometry (from R/02_get_ca_counties.R)
ca_counties_sf <- st_read(
  here::here("data", "processed", "boundaries", "ca_counties.gpkg"),
  quiet = TRUE
)

# Join totals to geometry
county_map <- ca_counties_sf %>%
  left_join(county_totals, by = "GEOID")

# Identify top 7 counties
top_n <- 7

top_counties <- county_totals %>%
  slice_max(total_obs, n = top_n, with_ties = FALSE)

top_counties_sf <- county_map %>%
  filter(GEOID %in% top_counties$GEOID) %>%
  mutate(centroid = st_centroid(geom))

# ---- Map ----
ggplot(county_map) +
  geom_sf(aes(fill = total_obs), color = "white", linewidth = 0.2) +
  
  geom_sf(
    data = top_counties_sf,
    fill = NA,
    color = "red",
    linewidth = 1.4
  ) +
  
  scale_fill_viridis(
    option = "magma",
    name = "Total observations"
  ) +
  
  labs(
    title = "Total GBIF bat observations by county (raw scale)",
    subtitle = "Most counties report very few or no observations; top 7 outlined in red"
  ) +
  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )


#Log-scaled choropleth map
ggplot(county_map) +
  geom_sf(aes(fill = total_obs), color = "white", linewidth = 0.2) +
  
  geom_sf(
    data = top_counties_sf,
    fill = NA,
    color = "red",
    linewidth = 1.4
  ) +
  
  scale_fill_viridis(
    option = "magma",
    trans = "log1p",
    name = "Total observations\n(log scale)"
  ) +
  
  labs(
    title = "Total GBIF bat observations by county (log-scaled)",
    subtitle = "Log(1 + observations) reveals variation among low-count counties"
  ) +
  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )




```

Bat observations are unevenly distributed across counties, with a small number of counties accounting for a disproportionate share of total observations. San Diego has the highest proportion of observations at ~27%. Only 4 counties account for 52% of bat observations: San Diego, Marin, Los Angeles, Humboldt. A small subset of counties contributes a majority of reported observations, indicating strong spatial concentration in reporting effort and/or bat activity. 


```{r}
#Counties with persistent zero counts ("data deserts")

county_zero_share <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    total_obs  = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))

county_zero_share

county_zero_share %>%
  filter(share_zero > 0.95)


```

18 counties have had 0 bat observations from 2013-2024. 27 counties have zero shares of ~97% or higher (only 1 total observation).
Several counties exhibit near-complete absence of reported observations across the study period, suggesting persistent gaps in reporting or unsuitable habitat rather than intermittent non-detection.



```{r}
#Spatial heterogeneity by species
#Do the same counties dominate across species?

data %>%
  group_by(county_name, species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(species_label, desc(total_obs))


ggplot(
  data %>%
    group_by(county_name, species_label) %>%
    summarise(total_obs = sum(n_obs), .groups = "drop"),
  aes(reorder(county_name, total_obs), total_obs)
) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Total observations by county and species",
    x = "County",
    y = "Total observations"
  )
```
Top counties
Overall:
1. San Diego	121		
2. Marin	57		
3. Los Angeles	40		
4. Humboldt	22		
5. San Francisco	20		
6. Santa Clara	16		
7. Sonoma 16

Aeroestes cinereus:
1. Los Angeles 29
2. San Diego 28
3. Marin 19
4. San Francisco 17
5. Humboldt 15
6. Santa Clara 15

Myotis cali
1. Marin	24		
2. San Diego	20
3. Napa 8		
4. Riverside	7		
5. Sonoma 7	

Myotis yuman
1. San Diego	73		
2. Marin	14		
3. Los Angeles	6		
4. Sonoma	6
5. Sutter 6

Top counties overall are San Diego (121), Marin (57), Los Angeles (40), Humboldt (22), San Francisco (20), Santa Clara (16), Sonoma (16)
San Diego and Marin are top counties across all 3 species.
A small set of coastal, urban, or research-adjacent counties account for a large fraction of all reported bat observations statewide. There is strog spatial concentration.
*Aeorestes cinereus observations are spatially widespread but concentrated in large, highly urbanized or coastal counties, suggesting strong interactions between migratory behavior, human observation effort, and landscape context.
*Myotis yumanensis is overwhelmingly driven by a single county (San Diego), whereas Myotis californicus shows moderate dominance by a small handful of counties.
*Super-counties are not identical across species, indicating that spatial dominance reflects species-specific ecology and detectability rather than a single statewide reporting bias.
Summary: These patterns indicate substantial county-level heterogeneity and species-specific spatial structure, motivating explicit consideration of county effects in subsequent analysis.

```{r}
#Distribution of counties by radiance level

county_radiance <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    mean_radiance = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(county_radiance, aes(mean_radiance)) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of mean nighttime radiance across counties",
    x = "Mean radiance (nW/cm²/sr)",
    y = "Number of counties"
  )

ggplot(county_radiance, aes(log1p(mean_radiance))) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of county radiance (log scale)",
    x = "log(1 + mean radiance)",
    y = "Number of counties"
  )


#Counties from highest to lowest radiance level
county_mean_radiance <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    mean_radiance_all_years = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_radiance_all_years))

county_mean_radiance
```
County-level mean nighttime radiance is extremely right-skewed. Roughly half of California counties exhibit near-zero mean radiance, while most remaining counties fall in a low radiance range (≈1–3). Only a small number of counties exhibit moderate to high radiance, with a very small extreme tail corresponding to highly urbanized regions.


```{r}
# ---- Maps: county radiance ----

# Join radiance onto county geometry
radiance_map <- ca_counties_sf %>%
  left_join(county_mean_radiance, by = "GEOID")

# Highlight top N brightest counties
top_n_bright <- 7

top_bright <- county_mean_radiance %>%
  slice_max(mean_radiance_all_years, n = top_n_bright, with_ties = FALSE)

top_bright_sf <- radiance_map %>%
  filter(GEOID %in% top_bright$GEOID)

# --- Map 1: RAW mean radiance ---
ggplot(radiance_map) +
  geom_sf(aes(fill = mean_radiance_all_years), color = "white", linewidth = 0.2) +
  geom_sf(data = top_bright_sf, fill = NA, color = "deepskyblue2", linewidth = 1.4) +
  scale_fill_viridis(
    option = "magma",
    name = "Mean radiance\n(all years)"
  ) +
  labs(
    title = "Mean nighttime radiance by California county (raw scale)",
    subtitle = paste0("Top ", top_n_bright, " brightest counties outlined")
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

# --- Map 2: LOG-SCALED mean radiance ---
ggplot(radiance_map) +
  geom_sf(aes(fill = mean_radiance_all_years), color = "white", linewidth = 0.2) +
  geom_sf(data = top_bright_sf, fill = NA, color = "deepskyblue2", linewidth = 1.4) +
  scale_fill_viridis(
    option = "magma",
    trans = "log1p",
    name = "Mean radiance\n(log1p scale)"
  ) +
  labs(
    title = "Mean nighttime radiance by California county (log-scaled)",
    subtitle = "Log(1 + radiance) reveals variation among low-radiance counties"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```



```{r}
#Are high-radiance counties also high-observation counties?

#raw radiance
county_summary <- county_totals %>%
  left_join(county_radiance, by = c("GEOID", "county_name"))

ggplot(county_summary, aes((mean_radiance), total_obs)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Total bat observations vs county radiance",
    x = "log(1 + mean radiance)",
    y = "Total observations"
  )

county_summary <- county_totals %>%
  left_join(county_radiance, by = c("GEOID", "county_name"))

#log radiance
ggplot(county_summary, aes(log1p(mean_radiance), total_obs)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Total bat observations vs county radiance",
    x = "log(1 + mean radiance)",
    y = "Total observations"
  )



```
The relationship between county-level nighttime radiance and total bat observations is weak and nonlinear. At low radiance values, there is a slight positive association, although many counties in this range report zero observations. At higher radiance levels, the relationship flattens, with many highly illuminated counties reporting few or no observations. A small number of counties (e.g., San Diego, Marin, Los Angeles) lie far above the overall trend, indicating that a limited set of super-counties disproportionately drive the observed association. 

Spatial exploratory analysis reveals strong heterogeneity in bat observations across California counties. A small number of counties contribute a disproportionate share of observations, while many exhibit persistent data sparsity. Nighttime radiance is also highly unevenly distributed, with most counties having very low light levels and a small subset of urban counties occupying the high-radiance extreme. The relationship between bat observations and light intensity is therefore strongly shaped by county-level factors, including observation effort and species-specific spatial patterns. These findings indicate that spatial heterogeneity must be carefully considered when interpreting bat–light associations.

***Any observed association between bat observations and nighttime light is shaped by where observations occur, not just how bright those places are.


### Light- VIIRS Structure

```{r}
#Aggregate mean radiance across all counties by year

data %>%
  group_by(year) %>%
  summarise(
    mean_radiance_state = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, mean_radiance_state)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Statewide mean nighttime radiance over time",
    x = "Year",
    y = "Mean radiance (nW/cm²/sr)"
  )
```
~12–13% relative variation, spread across more than a decade, with no monotonic trend
Statewide mean nighttime radiance fluctuates modestly from year to year but remains within a narrow range across the study period. These changes are small relative to the much larger differences observed between counties, indicating that radiance is a largely stable spatial characteristic. 
```{r}
#County-level radiance trends
#Are changes in light driven by all counties or just a few?

top_bright_ids <- county_mean_radiance %>%
  slice_max(mean_radiance_all_years, n = 10) %>%
  pull(GEOID)

data %>%
  filter(GEOID %in% top_bright_ids) %>%
  ggplot(aes(year, mean_radiance, group = county_name, color = county_name)) +
  geom_line() +
  labs(
    title = "Nighttime radiance trends in brightest counties",
    x = "Year",
    y = "Mean radiance"
  ) +
  theme(legend.position = "none")

```
Year-to-year variation is driven primarily by a small number of very bright counties, while most counties exhibit minimal temporal change.

```{r}
# Between-county variance (using long-term means)
var_between <- county_mean_radiance %>%
  summarise(var_between = var(mean_radiance_all_years)) %>%
  pull(var_between)

# Within-county variance (year-to-year)
var_within <- data %>%
  group_by(GEOID) %>%
  summarise(var_within = var(mean_radiance, na.rm = TRUE)) %>%
  summarise(mean_var_within = mean(var_within, na.rm = TRUE)) %>%
  pull(mean_var_within)

var_between
var_within
```
Between-county variance >> within-county variance

Variation in nighttime light is dominated by differences between counties rather than changes within counties over time, indicating that radiance primarily captures persistent spatial context rather than temporal dynamics.
This justifies aggregating over years.



```{r}
#Radiance distribution within species’ observed counties

data %>%
  filter(n_obs > 0) %>%
  ggplot(aes((mean_radiance))) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ species_label) +
  labs(
    title = "Radiance distribution in counties with bat observations",
    x = "Mean Radiance",
    y = "Count"
  )

data %>%
  filter(n_obs > 0) %>%
  ggplot(aes(log1p(mean_radiance))) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ species_label) +
  labs(
    title = "Log Radiance distribution in counties with bat observations",
    x = "log(1 + mean radiance)",
    y = "Count"
  )
```
Across all 3 species, the most observations take place in the darker counties of less than 5 in mean radiance. Close to 0 in mean radiance, there are almost twice as many Aeorestes observed than Mytosis californicus and twice as many Aeroestes as Myotis yumanensis. Aeroestes also has a wider spread across different light intensities, with more observations found in counties above 20 radiance and even 30 radiance than both Myotis species combined.

```{r}
#Relationship between radiance and observation coverage
#Are brighter counties more likely to have any bat observations?
data %>%
  mutate(
    radiance_bin = cut(
      mean_radiance,
      breaks = c(0, 0.1, 1, 3, 10, Inf),
      include.lowest = TRUE
    )
  ) %>%
  group_by(radiance_bin) %>%
  summarise(
    share_nonzero = mean(n_obs > 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(radiance_bin, share_nonzero)) +
  geom_col() +
  labs(
    title = "Share of county-years with bat observations by radiance level",
    x = "Mean radiance bin",
    y = "Proportion with ≥1 observation"
  )
```
Brighter counties are much more likely to have at least one reported bat observation in a given year. Bright counties don't dominate counts but they do dominate coverage. Nighttime light is strongly associated with the probability of observation, but not necessarily with observation intensity. Light helps explain where bats are observed at all. Light does not explain how many bats are observed once present


Summary: Exploratory analysis of VIIRS nighttime light data indicates that radiance is overwhelmingly dominated by spatial rather than temporal variation, with between-county differences far exceeding within-county year-to-year changes. Across all three focal species, most bat observations occur in counties with low mean radiance, indicating that high light intensity is not a prerequisite for reported bat activity. At the same time, the likelihood of observing bats at least once in a given county–year increases with nighttime radiance, suggesting that light primarily influences observation coverage rather than observation intensity. However, strong species-specific differences are evident: Aeorestes cinereus exhibits a much broader distribution across the radiance gradient, including frequent observations in highly illuminated counties, while Myotis species are largely confined to darker environments. Together, these patterns suggest that nighttime light captures long-term spatial context and interacts with species ecology and observation processes, motivating species-aware and spatially cautious interpretation in subsequent analyses.


### Zero Structure vs Light

```{r}
#Create a zero indicator

data <- data %>%
  mutate(is_zero = (n_obs == 0))

#Share of zero rows by radiance decile
zero_by_decile <- data %>%
  mutate(
    radiance_decile = ntile(log1p_radiance, 10)
  ) %>%
  group_by(radiance_decile) %>%
  summarise(
    n_rows = n(),
    share_zero = mean(is_zero),
    share_nonzero = mean(!is_zero),
    .groups = "drop"
  )

zero_by_decile

ggplot(zero_by_decile, aes(x = radiance_decile, y = share_zero)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows by radiance decile",
    subtitle = "Deciles computed from log1p_radiance across all county-years and species",
    x = "Radiance decile (1 = lowest, 10 = highest)",
    y = "Share of rows with n_obs = 0"
  )


#By species
zero_by_decile_species <- data %>%
  mutate(
    radiance_decile = ntile(log1p_radiance, 10)
  ) %>%
  group_by(species_label, radiance_decile) %>%
  summarise(
    n_rows = n(),
    share_zero = mean(is_zero),
    .groups = "drop"
  )

zero_by_decile_species

ggplot(zero_by_decile_species, aes(radiance_decile, share_zero, color = species_label)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows by radiance decile (by species)",
    x = "Radiance decile (1 = lowest, 10 = highest)",
    y = "Share of rows with n_obs = 0",
    color = "Species"
  )
```
Zeroes are more common at low radiance , although it is not a monotonically negative association. But even in the brightest counties, most rows are still zero. All 3 species show similar trends, with Aeorestes having the strongest reaction to increasing radiance. Light strongly reduces zeros for Aeorestes, but only weakly for Myotis species.

```{r}
#Share of zero rows by county radiance rank

#Create county radiance ranks
county_radiance_rank <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    county_mean_log_radiance = mean(log1p_radiance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    radiance_rank = percent_rank(county_mean_log_radiance),  # 0=darkest, 1=brightest
    radiance_quintile = ntile(county_mean_log_radiance, 5)
  )

county_radiance_rank


#Join ranks back + compute zero share by quintile
zero_by_county_quintile <- data %>%
  left_join(county_radiance_rank %>% select(GEOID, radiance_quintile),
            by = "GEOID") %>%
  group_by(radiance_quintile) %>%
  summarise(
    n_rows = n(),
    share_zero = mean(is_zero),
    .groups = "drop"
  )

zero_by_county_quintile

ggplot(zero_by_county_quintile, aes(factor(radiance_quintile), share_zero)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows by county radiance quintile",
    subtitle = "Counties ranked by their long-term mean log radiance",
    x = "County radiance quintile (1 = darkest, 5 = brightest)",
    y = "Share of rows with n_obs = 0"
  )


zero_by_county_quintile_species <- data %>%
  left_join(county_radiance_rank %>% select(GEOID, radiance_quintile),
            by = "GEOID") %>%
  group_by(species_label, radiance_quintile) %>%
  summarise(
    n_rows = n(),
    share_zero = mean(is_zero),
    .groups = "drop"
  )

#Same thing by species
ggplot(zero_by_county_quintile_species, aes(factor(radiance_quintile), share_zero, fill = species_label)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows by county radiance quintile (by species)",
    x = "County radiance quintile (1 = darkest, 5 = brightest)",
    y = "Share of rows with n_obs = 0",
    fill = "Species"
  )



```
The darkest counties have about 96% zero rows while the brightest counties have 81% zero rows.
Bright counties are much more likely to ever have observations, but they still produce many zero years/species combinations.

Aeorestes has a dramatic drop in zeros in bright counties while Myotis has much weaker dependence on county brightness. Species differences dominate the zero process as much as, or more than, light itself.



```{r}
#Zero probability vs radiance curves

data <- data %>%
  mutate(
    is_zero = if_else(n_obs == 0, 1, 0)) #Make is_zero a numeric variable

#Overall curve
ggplot(data, aes(log1p_radiance, is_zero)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = TRUE) +
  labs(
    title = "Estimated probability of zero observations vs radiance",
    subtitle = "Binomial smoother: Pr(n_obs = 0) as a function of log1p_radiance",
    x = "log(1 + mean radiance)",
    y = "Pr(n_obs = 0)"
  )

#Species specific curves
ggplot(data, aes(log1p_radiance, is_zero, color = species_label)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = FALSE) +
  labs(
    title = "Zero probability vs radiance (by species)",
    subtitle = "Binomial smoother: Pr(n_obs = 0)",
    x = "log(1 + mean radiance)",
    y = "Pr(n_obs = 0)",
    color = "Species"
  )

ggplot(data, aes(log1p_radiance, is_zero)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = TRUE) +
  facet_wrap(~ species_label) +
  labs(
    title = "Zero probability vs radiance (faceted by species)",
    x = "log(1 + mean radiance)",
    y = "Pr(n_obs = 0)"
  )
```

The pooled curve shows a smooth, monotonic decline that confirms a continuous relationship. Therefore a log transform is appropriate.

The species-specific curves: Aeroestes has a steep, non-linear decline in zero probability. Myotis species are shallow nearly parallel curves with high zero probability.

Conclusions:
1) Zeros are more common in dark counties, less common in bright counties, and strongly species-dependent.
2) Nighttime light affects presence, not intensity. Light reduces probability of a zero but does not guarantee high counts. Many bright counties still have zero observations.
3) Aeroestes drives most of the light-zero relationship.
4) Zero inflation is real and unavoidable.

### Species Specific Behavior


```{r}

#Build county-species summary

# County-species totals (sum over years)
county_species_totals <- data %>%
  group_by(GEOID, county_name, species_label) %>%
  summarise(
    total_obs = sum(n_obs, na.rm = TRUE),
    .groups = "drop"
  )

# County-species radiance (mean over years)
county_species_radiance <- data %>%
  group_by(GEOID, county_name, species_label) %>%
  summarise(
    mean_radiance = mean(mean_radiance, na.rm = TRUE),
    log1p_radiance = mean(log1p_radiance, na.rm = TRUE),  # you already have this per row; averaging is fine
    .groups = "drop"
  )

# Join into one dataset for plotting
county_species_summary <- county_species_totals %>%
  left_join(county_species_radiance, by = c("GEOID", "county_name", "species_label"))

```
```{r}
#Raw radiance by species
ggplot(county_species_summary, aes(mean_radiance, total_obs, color = species_label)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Total bat observations vs county radiance (all species)",
    subtitle = "Each point is a county × species; totals summed over years",
    x = "Mean radiance (nW/cm²/sr)",
    y = "Total observations (2012–2024)",
    color = "Species"
  )

#Log radiance by species
ggplot(county_species_summary, aes(log1p_radiance, total_obs, color = species_label)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Total bat observations vs county radiance (log scale, all species)",
    subtitle = "Each point is a county × species; totals summed over years",
    x = "log(1 + mean radiance)",
    y = "Total observations (2012–2024)",
    color = "Species"
  )
```

```{r}
#Faceted plot by species

#Raw radiance
ggplot(county_species_summary, aes(mean_radiance, total_obs)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Total bat observations vs county radiance (by species)",
    subtitle = "Each point is a county; totals summed over years",
    x = "Mean radiance (nW/cm²/sr)",
    y = "Total observations (2012–2024)"
  )

#LOG radiance
ggplot(county_species_summary, aes(log1p_radiance, total_obs)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Total bat observations vs county radiance (log scale, by species)",
    subtitle = "Each point is a county; radiance shown as log(1 + radiance)",
    x = "log(1 + mean radiance)",
    y = "Total observations (2012–2024)"
  )


```
Conclusions:
1) There is no single bat-light relationship.
2) Log-radiance is the correct scale
3) Aeorestes cinereus shows strong, monotonic increase in total observations with increasing radiance. Aeorestes is the only species with:
*substantial totals in very bright counties
*a rising loess curve at high radiance 
Aeorestes is either:
*more tolerant of human-modified, illuminated environments, and/or
*more likely to be detected and reported in urban counties
4) Myotis californicus shows a small peak at low-to-moderate radiance and then declining totals as radiance decreases. This species appears constrained to darker environments, suggesting habitat specialization, avoidance of highly illuminated landscape, or low detectability in urban contexts.
5) Myotis yumanensis has very spatially concentrated observations (San Diego is extreme outlier) and is rare everywhere else regardless of light level. This reinforces sensitivity to spatial context, importance of super-counties, and the limits of light as a predictor.
6) Observed pattern reflect a mix of ecology and detection bias.

### County trajectories 9within-county change)

```{r}
#Small informative set of counties
focus_counties <- c(
  "San Diego",
  "Los Angeles",
  "Marin",
  "Humboldt",
  "Sutter"
)

focus_data <- data %>%
  filter(county_name %in% focus_counties)

#County-level radiance time series
focus_data %>%
  group_by(county_name, year) %>%
  summarise(
    mean_radiance = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, mean_radiance, color = county_name)) +
  geom_line() +
  geom_point() +
  labs(
    title = "County-level nighttime radiance over time",
    x = "Year",
    y = "Mean radiance (nW/cm²/sr)",
    color = "County"
  )

#County-level observation time series
focus_data %>%
  group_by(county_name, species_label, year) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs, color = species_label)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ county_name, scales = "free_y") +
  labs(
    title = "Bat observations over time by county and species",
    x = "Year",
    y = "Total observations",
    color = "Species"
  )


```

```{r}
#When radiance changes within a county, do observations change in the opposite direction?
focus_data %>%
  group_by(county_name, species_label) %>%
  arrange(year) %>%
  mutate(
    delta_radiance = mean_radiance - lag(mean_radiance),
    delta_obs = n_obs - lag(n_obs)
  ) %>%
  filter(!is.na(delta_radiance), !is.na(delta_obs)) %>%
  ggplot(aes(delta_radiance, delta_obs)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ species_label) +
  labs(
    title = "Within-county year-to-year changes",
    subtitle = "Δ observations vs Δ radiance",
    x = "Change in mean radiance",
    y = "Change in observations"
  )

```
Nighttime light is essentially a fixed spatial characteristic at the county scale over this time period. Temporal variation in bat observations appears driven by sporadic events and reporting effort, not by gradual environmental change. There is little evidence of within-county temporal coupling between changes in nighttime light and changes in bat observations.

Summary: Examination of within-county trajectories shows that nighttime radiance is highly stable over time, with only minor year-to-year variation relative to persistent differences between counties. In contrast, bat observation counts within counties are highly variable and dominated by intermittent spikes rather than consistent trends. Analyses of year-to-year changes reveal no systematic relationship between changes in radiance and changes in bat observations, either contemporaneously or with a lag. These findings indicate that bat–light relationships in this dataset are primarily spatial rather than temporal in nature, supporting aggregation across years and motivating spatially structured or fixed-effects modeling approaches rather than longitudinal causal analysis.

