---
title: "eda-sandbox"
output: html_notebook
---

### Distribution of the Response (n_obs)

```{r}
#Load Packages and Data

library(tidyverse)
library(janitor)
library(ggplot2)


data <- readr::read_csv(
  here::here(
    "data/processed/analysis/bats_viirs_county_year_species.csv"
  ),
  show_col_types = TRUE
)
```
```{r}
#Check out data

glimpse(data)

summary(dat$n_obs)

```

All the zeroes bring the means number of observations per species-year-county down to only 0.2. Max number of observations in any one row is 19.

```{r}
#Proportion of zeros overall

data %>%
  summarise(
    share_zero = mean(n_obs == 0),
    share_nonzero = mean(n_obs > 0)
  )

#Counts of each species

data %>%
  group_by(species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  )

#Zeros by species

data %>%
  group_by(species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    mean_n_obs = mean(n_obs),
    max_n_obs  = max(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))
```

About 90% of all rows have 0 observations, indicating sparsity and zero inflation.
Aeorestes (hoary bat) has about 85% share of 0 observations while Myotis has 93-94%. This could say something about the difference in ease of detecting those species.

The majority of species–county–year combinations contain zero reported observations. This sparsity reflects the opportunistic nature of GBIF records, coarse spatial aggregation, and imperfect detection of nocturnal bats rather than true absence. As a result, observation counts are interpreted as relative reporting intensity rather than abundance.

```{r}
#Histogram of n_obs (raw scale)
ggplot(dat, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  labs(
    title = "Distribution of bat observation counts",
    x = "Number of observations (n_obs)",
    y = "Number of species–county–year rows"
  )

#Histogram on log scale (to see tail structure)
ggplot(dat %>% filter(n_obs > 0), aes(n_obs)) +
  geom_histogram(binwidth = 1, color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of nonzero bat observations (log scale)",
    x = "n_obs (log scale)",
    y = "Count"
  )


#Variance vs mean
dat %>%
  summarise(
    mean_n_obs = mean(n_obs),
    var_n_obs  = var(n_obs)
  )


```
The distribution of nonzero observation counts is heavily right-skewed (small but influential tail!). Most species–county–year combinations with reported observations have only 1–3 records, while a small number exhibit substantially higher counts (up to ~30). This long-tailed structure indicates strong overdispersion and suggests that a small subset of counties and years account for a disproportionate share of reported bat activity.

Also, mean observations are 0.20 and variance of observations is 0.89. Variance > mean so that is overdispersion.Te small tail dominates the variance


```{r}
#Response distribution by species (faceted)
ggplot(dat, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Distribution of observation counts by species",
    x = "n_obs",
    y = "Count"
  )
```

Observation counts are highly overdispersed relative to a Poisson distribution (mean ≈ 0.2, variance ≈ 0.89). While most nonzero counts are small (1–3 observations), a small number of county–year–species combinations exhibit moderately large counts, producing a long right tail. This pattern is most pronounced for Aeorestes cinereus, which shows both fewer zeros and higher maximum counts than the Myotis species.

The Poisson distribution is inappropriate for this dataset because observation counts are highly overdispersed relative to the mean, contain excess zeros, and reflect unmodeled heterogeneity in observation effort and habitat suitability across counties, years, and species.


### Temporal Structure

```{r}
#Total observations per year (overall)

data %>%
  group_by(year) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total reported bat observations per year",
    x = "Year",
    y = "Total observations"
  )

#Observations per year by species

dat %>%
  group_by(year, species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total bat observations per year by species",
    x = "Year",
    y = "Total observations",
    color = "Species"
  )

```
I wonder what caused the peak in 2016?
In the earlier years, Myotis observations outnumber Aeroestes but by 2020, Aeroestes observations are much higher.




```{r}
#Number of counties with ≥1 observation per year
data %>%
  group_by(year) %>%
  summarise(
    n_counties_with_obs = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties_with_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of counties with ≥1 bat observation per year",
    x = "Year",
    y = "Counties reporting bats"
  )

#Species-specific county coverage over time
data %>%
  group_by(year, species_label) %>%
  summarise(
    n_counties = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "County coverage by species over time",
    x = "Year",
    y = "Counties with ≥1 observation",
    color = "Species"
  )
```
There  is a lot of fluctuation from year to year but overall appears to be a general trend towards spatial expansion of bat reporting.
Aeroestes had an increase in how widely they're reported over time, while the Myotis species both decreased, meaning the overall spatial expansion of bat reporting was driven by Aeroestes


```{r}
#Zero share over time- are zeros becoming less common?

data %>%
  group_by(year) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows over time",
    x = "Year",
    y = "Percent zero"
  )

#Zero share by species over time

data %>%
  group_by(year, species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero, color = species_label)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Zero-observation share over time by species",
    x = "Year",
    y = "Percent zero",
    color = "Species"
  )


```
Share of zeros over time has had a few percentage jumps but seems to have stabled around 90% in recent years.

Summary: Bat observation data exhibit strong temporal heterogeneity rather than a simple trend. Total observations peak sharply in 2016, driven primarily by Myotis californicus, followed by a decline and lower-level fluctuations in subsequent years. In contrast, Aeorestes cinereus observations become increasingly dominant after 2017, accounting for a growing share of both total observations and spatial coverage. While total counts decline after 2016, the number of counties with at least one reported observation increases over time, indicating broader but shallower reporting in recent years. High proportions of zero-observation rows persist throughout the study period, particularly for Myotis species, highlighting structural sparsity and species-specific detection dynamics.



### Spatial Structure


```{r}
#Total observations by county (summed over years)

county_totals <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(total_obs))

county_totals


ggplot(county_totals, aes(reorder(county_name, total_obs), total_obs)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Total bat observations by county (2012–2024)",
    x = "County",
    y = "Total observations"
  )


#Counties dominating total observaitons ("super-observer" counties)
#Cumulative contribution
county_totals %>%
  mutate(
    cumulative_obs = cumsum(total_obs),
    cumulative_share = cumulative_obs / sum(total_obs)
  )




```

Bat observations are unevenly distributed across counties, with a small number of counties accounting for a disproportionate share of total observations. San Diego has the highest proportion of observations at ~27%. Only 4 counties account for 52% of bat observations: San Diego, Marin, Los Angeles, Humboldt. A small subset of counties contributes a majority of reported observations, indicating strong spatial concentration in reporting effort and/or bat activity. 


```{r}
#Counties with persistent zero counts

county_zero_share <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    total_obs  = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))

county_zero_share

county_zero_share %>%
  filter(share_zero > 0.95)


```

18 counties have had 0 bat observations from 2013-2024. 27 counties have zero shares of ~97% or higher (only 1 total observation).
Several counties exhibit near-complete absence of reported observations across the study period, suggesting persistent gaps in reporting or unsuitable habitat rather than intermittent non-detection.



```{r}
#Spatial heterogeneity by species
#Do the same counties dominate across species?

data %>%
  group_by(county_name, species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  arrange(species_label, desc(total_obs))


ggplot(
  data %>%
    group_by(county_name, species_label) %>%
    summarise(total_obs = sum(n_obs), .groups = "drop"),
  aes(reorder(county_name, total_obs), total_obs)
) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Total observations by county and species",
    x = "County",
    y = "Total observations"
  )
```
Top counties
Overall:
1. San Diego	121		
2. Marin	57		
3. Los Angeles	40		
4. Humboldt	22		
5. San Francisco	20		
6. Santa Clara	16		
7. Sonoma 16

Aeroestes cinereus:
1. Los Angeles 29
2. San Diego 28
3. Marin 19
4. San Francisco 17
5. Humboldt 15
6. Santa Clara 15

Myotis cali
1. Marin	24		
2. San Diego	20
3. Napa 8		
4. Riverside	7		
5. Sonoma 7	

Myotis yuman
1. San Diego	73		
2. Marin	14		
3. Los Angeles	6		
4. Sonoma	6
5. Sutter 6

Top counties overall are San Diego (121), Marin (57), Los Angeles (40), Humboldt (22), San Francisco (20), Santa Clara (16), Sonoma (16)
San Diego and Marin are top counties across all 3 species.
A small set of coastal, urban, or research-adjacent counties account for a large fraction of all reported bat observations statewide. There is strog spatial concentration.
*Aeorestes cinereus observations are spatially widespread but concentrated in large, highly urbanized or coastal counties, suggesting strong interactions between migratory behavior, human observation effort, and landscape context.
*Myotis yumanensis is overwhelmingly driven by a single county (San Diego), whereas Myotis californicus shows moderate dominance by a small handful of counties.
*Super-counties are not identical across species, indicating that spatial dominance reflects species-specific ecology and detectability rather than a single statewide reporting bias.
Summary: These patterns indicate substantial county-level heterogeneity and species-specific spatial structure, motivating explicit consideration of county effects in subsequent analysis.

```{r}
#Distribution of counties by radiance level

county_radiance <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    mean_radiance = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(county_radiance, aes(mean_radiance)) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of mean nighttime radiance across counties",
    x = "Mean radiance (nW/cm²/sr)",
    y = "Number of counties"
  )

ggplot(county_radiance, aes(log1p(mean_radiance))) +
  geom_histogram(bins = 30, color = "white") +
  labs(
    title = "Distribution of county radiance (log scale)",
    x = "log(1 + mean radiance)",
    y = "Number of counties"
  )


#Counties from highest to lowest radiance level
county_mean_radiance <- data %>%
  group_by(GEOID, county_name) %>%
  summarise(
    mean_radiance_all_years = mean(mean_radiance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_radiance_all_years))

county_mean_radiance
```
County-level mean nighttime radiance is extremely right-skewed. Roughly half of California counties exhibit near-zero mean radiance, while most remaining counties fall in a low radiance range (≈1–3). Only a small number of counties exhibit moderate to high radiance, with a very small extreme tail corresponding to highly urbanized regions.



```{r}
#Are high-radiance counties also high-observation counties?
county_summary <- county_totals %>%
  left_join(county_radiance, by = c("GEOID", "county_name"))

ggplot(county_summary, aes(log1p(mean_radiance), total_obs)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Total bat observations vs county radiance",
    x = "log(1 + mean radiance)",
    y = "Total observations"
  )



```
The relationship between county-level nighttime radiance and total bat observations is weak and nonlinear. At low radiance values, there is a slight positive association, although many counties in this range report zero observations. At higher radiance levels, the relationship flattens, with many highly illuminated counties reporting few or no observations. A small number of counties (e.g., San Diego, Marin, Los Angeles) lie far above the overall trend, indicating that a limited set of super-counties disproportionately drive the observed association. 

Spatial exploratory analysis reveals strong heterogeneity in bat observations across California counties. A small number of counties contribute a disproportionate share of observations, while many exhibit persistent data sparsity. Nighttime radiance is also highly unevenly distributed, with most counties having very low light levels and a small subset of urban counties occupying the high-radiance extreme. The relationship between bat observations and light intensity is therefore strongly shaped by county-level factors, including observation effort and species-specific spatial patterns. These findings indicate that spatial heterogeneity must be carefully considered when interpreting bat–light associations.

***Any observed association between bat observations and nighttime light is shaped by where observations occur, not just how bright those places are.






































