---
title: "eda-sandbox"
output: html_notebook
---

### Distribution of the Response (n_obs)

```{r}
#Load Packages and Data

library(tidyverse)
library(janitor)
library(ggplot2)


data <- readr::read_csv(
  here::here(
    "data/processed/analysis/bats_viirs_county_year_species.csv"
  ),
  show_col_types = TRUE
)
```
```{r}
#Check out data

glimpse(data)

summary(dat$n_obs)

```

All the zeroes bring the means number of observations per species-year-county down to only 0.2. Max number of observations in any one row is 19.

```{r}
#Proportion of zeros overall

data %>%
  summarise(
    share_zero = mean(n_obs == 0),
    share_nonzero = mean(n_obs > 0)
  )

#Counts of each species

data %>%
  group_by(species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  )

#Zeros by species

data %>%
  group_by(species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    mean_n_obs = mean(n_obs),
    max_n_obs  = max(n_obs),
    .groups = "drop"
  ) %>%
  arrange(desc(share_zero))
```

About 90% of all rows have 0 observations, indicating sparsity and zero inflation.
Aeorestes (hoary bat) has about 85% share of 0 observations while Myotis has 93-94%. This could say something about the difference in ease of detecting those species.

The majority of species–county–year combinations contain zero reported observations. This sparsity reflects the opportunistic nature of GBIF records, coarse spatial aggregation, and imperfect detection of nocturnal bats rather than true absence. As a result, observation counts are interpreted as relative reporting intensity rather than abundance.

```{r}
#Histogram of n_obs (raw scale)
ggplot(dat, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  labs(
    title = "Distribution of bat observation counts",
    x = "Number of observations (n_obs)",
    y = "Number of species–county–year rows"
  )

#Histogram on log scale (to see tail structure)
ggplot(dat %>% filter(n_obs > 0), aes(n_obs)) +
  geom_histogram(binwidth = 1, color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of nonzero bat observations (log scale)",
    x = "n_obs (log scale)",
    y = "Count"
  )


#Variance vs mean
dat %>%
  summarise(
    mean_n_obs = mean(n_obs),
    var_n_obs  = var(n_obs)
  )


```
The distribution of nonzero observation counts is heavily right-skewed (small but influential tail!). Most species–county–year combinations with reported observations have only 1–3 records, while a small number exhibit substantially higher counts (up to ~30). This long-tailed structure indicates strong overdispersion and suggests that a small subset of counties and years account for a disproportionate share of reported bat activity.

Also, mean observations are 0.20 and variance of observations is 0.89. Variance > mean so that is overdispersion.Te small tail dominates the variance


```{r}
#Response distribution by species (faceted)
ggplot(dat, aes(n_obs)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white") +
  facet_wrap(~ species_label, scales = "free_y") +
  labs(
    title = "Distribution of observation counts by species",
    x = "n_obs",
    y = "Count"
  )
```

Observation counts are highly overdispersed relative to a Poisson distribution (mean ≈ 0.2, variance ≈ 0.89). While most nonzero counts are small (1–3 observations), a small number of county–year–species combinations exhibit moderately large counts, producing a long right tail. This pattern is most pronounced for Aeorestes cinereus, which shows both fewer zeros and higher maximum counts than the Myotis species.

The Poisson distribution is inappropriate for this dataset because observation counts are highly overdispersed relative to the mean, contain excess zeros, and reflect unmodeled heterogeneity in observation effort and habitat suitability across counties, years, and species.


### Temporal Structure

```{r}
#Total observations per year (overall)

data %>%
  group_by(year) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total reported bat observations per year",
    x = "Year",
    y = "Total observations"
  )

#Observations per year by species

dat %>%
  group_by(year, species_label) %>%
  summarise(
    total_obs = sum(n_obs),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, total_obs, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total bat observations per year by species",
    x = "Year",
    y = "Total observations",
    color = "Species"
  )

```
I wonder what caused the peak in 2016?
In the earlier years, Myotis observations outnumber Aeroestes but by 2020, Aeroestes observations are much higher.




```{r}
#Number of counties with ≥1 observation per year
data %>%
  group_by(year) %>%
  summarise(
    n_counties_with_obs = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties_with_obs)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of counties with ≥1 bat observation per year",
    x = "Year",
    y = "Counties reporting bats"
  )

#Species-specific county coverage over time
data %>%
  group_by(year, species_label) %>%
  summarise(
    n_counties = n_distinct(GEOID[n_obs > 0]),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, n_counties, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "County coverage by species over time",
    x = "Year",
    y = "Counties with ≥1 observation",
    color = "Species"
  )
```
There  is a lot of fluctuation from year to year but overall appears to be a general trend towards spatial expansion of bat reporting.
Aeroestes had an increase in how widely they're reported over time, while the Myotis species both decreased, meaning the overall spatial expansion of bat reporting was driven by Aeroestes


```{r}
#Zero share over time- are zeros becoming less common?

data %>%
  group_by(year) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of zero-observation rows over time",
    x = "Year",
    y = "Percent zero"
  )

#Zero share by species over time

data %>%
  group_by(year, species_label) %>%
  summarise(
    share_zero = mean(n_obs == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(year, share_zero, color = species_label)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Zero-observation share over time by species",
    x = "Year",
    y = "Percent zero",
    color = "Species"
  )


```
Share of zeros over time has had a few percentage jumps but seems to have stabled around 90% in recent years.

Summary: Bat observation data exhibit strong temporal heterogeneity rather than a simple trend. Total observations peak sharply in 2016, driven primarily by Myotis californicus, followed by a decline and lower-level fluctuations in subsequent years. In contrast, Aeorestes cinereus observations become increasingly dominant after 2017, accounting for a growing share of both total observations and spatial coverage. While total counts decline after 2016, the number of counties with at least one reported observation increases over time, indicating broader but shallower reporting in recent years. High proportions of zero-observation rows persist throughout the study period, particularly for Myotis species, highlighting structural sparsity and species-specific detection dynamics.



### Spatial Structure


```{r}

```













































